\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amsthm, amssymb, amsfonts}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{enumitem}
\usepackage{natbib}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue!50!black, citecolor=blue!50!black, urlcolor=blue!50!black}

% Theorem environments
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

% Shortcuts
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\abs}[1]{\left\lvert#1\right\rvert}
\newcommand{\inner}[2]{\left\langle#1,#2\right\rangle}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\trace}{tr}
\DeclareMathOperator{\spec}{spec}
\newcommand{\method}{polynomial input preconditioning}
\newcommand{\Method}{Polynomial input preconditioning}
\newcommand{\HD}{\textsf{d4\_dropout}}
\newcommand{\Dnodrop}{\textsf{d4}}
\newcommand{\Baseline}{\textsf{Baseline}}
\newcommand{\Zero}{\textsf{Zero}}
\newcommand{\Dup}{\textsf{Duplicate}}

\title{\textbf{Polynomial Input Preconditioning for\\Universal Time Series Forecasting}}
\author{
  Jerry Han \\[0.5em]
  \small Department of Mathematics \\
  \small Princeton University \\[0.5em]
  \small Advisor: Prof.\ Elad Hazan \\
  \small Second Reader: Prof.\ Amit Singer
}
\date{Spring 2026}

\begin{document}
\maketitle

\begin{abstract}
\noindent
\emph{Universal Sequence Preconditioning} (USP) is a recent theoretical framework~\citep{marsden2024universal} for online sequence prediction in which a fixed convolutional transformation of the target sequence---specifically, convolution with the coefficients of a monic Chebyshev or Legendre polynomial---yields provable, hidden-dimension-independent regret bounds for marginally stable linear dynamical systems (LDS). This paper investigates whether the USP idea transfers, empirically and theoretically, to a setting that the original framework does not directly cover: \emph{patch-based transformer foundation models for time-series forecasting}, such as Moirai-2~\citep{liu2024moirai2}. The setting differs from USP in three ways---the forecaster is a patch-based transformer rather than an online learner, the data may be far richer than a finite-dimensional LDS, and the forecasting task is multi-step rather than single-step. To accommodate these differences we propose \textbf{\method}, in which the polynomial-preconditioned signal is concatenated as an auxiliary input \emph{channel} alongside the raw target rather than replacing it, sidestepping the catastrophic error accumulation that pure target preconditioning would induce under autoregressive multi-step rollout. Architecturally, only the per-patch input projection is widened; the transformer decoder is unchanged, adding $12$K parameters or $0.11\%$ of the Moirai-2 Small model. We give a self-contained mathematical treatment grounded in the Chebyshev equioscillation theorem and a closed-form analysis of our filter's frequency response, a noise-gain analysis with a spectral--noise tradeoff, and a comparison of the Chebyshev basis with Legendre and second-kind alternatives. Empirically, on the GIFT-Eval benchmark ($97$ dataset $\times$ horizon configurations, 5 seeds, $485$ paired comparisons) our method achieves a $\Delta\mathrm{MASE}^* = -2.9\%$ relative improvement in the normalised geometric-mean MASE over a from-scratch Moirai-2 Small baseline, winning $329$ of $485$ paired per-configuration comparisons; capacity-matched controls (duplicate channel, zero channel) confirm that the polynomial content---not the extra parameters---drives the gain. At $100$K training steps the improvement grows to $\Delta\mathrm{MASE}^* = -3.9\%$, and the gain remains around $-5\%$ across context lengths $512$--$4{,}000$. Throughout the paper we report relative MASE changes with the convention $\Delta\% = (\text{treatment} - \text{control})/\text{control}$, so a negative $\Delta\%$ means lower MASE (an improvement).
\end{abstract}

\section*{Acknowledgments and Contributions}

I am deeply grateful to my advisor, Prof.\ Elad Hazan, for suggesting the connection between universal sequence preconditioning and patch-based time series foundation models, and for numerous conversations that shaped the direction of this work. I thank Prof.\ Amit Singer for serving as second reader and for feedback on the mathematical exposition. Any errors are my own.

\paragraph{Individual contributions.}
The theoretical formulation, experimental design, implementation, and writing presented in this paper are my individual work. The framework of universal sequence preconditioning as applied to online prediction of linear dynamical systems is due to \citet{marsden2024universal}; the extension of that framework to patch-based transformer forecasting, the specific filter design, the empirical methodology, the statistical analysis, and all figures and tables were developed by me. Computational experiments were run on the Princeton Research Computing cluster.

\paragraph{Relation to workshop submission.}
An abridged 4-page version of the empirical portion of this work was submitted concurrently to the ICML 2026 Workshop on Foundation Models for Structured Data (FMSD); the workshop is non-archival. This Junior Paper contains substantially expanded mathematical analysis and additional theoretical results not present in the workshop submission.

\paragraph{Honor Code Declaration.}
This paper represents my own work in accordance with University regulations.\\[0.5em]
\textit{/s/ Jerry Han}

\tableofcontents

%==============================================================================
\section{Introduction}
%==============================================================================

\paragraph{Universal sequence preconditioning.}
\citet{marsden2024universal} recently proved provable guarantees for online prediction algorithms on data arising from marginally stable, possibly asymmetric, linear dynamical systems (LDS). Their key observation: for an LDS with hidden-state recurrence $x_{t+1} = A x_t + w_t$ and observation $y_t = C x_t + v_t$, convolving the target sequence with the coefficients of a degree-$d$ monic polynomial $p(z)$ is algebraically equivalent to multiplying the hidden dynamics by $p(A)$. Choosing $p$ to be the monic Chebyshev polynomial of the first kind, $\tilde T_d$, exploits its classical equioscillation property: among all monic degree-$d$ polynomials, $\tilde T_d$ uniquely attains the minimal $\ell_\infty$ norm $2^{1-d}$ on $[-1,1]$~\citep{davis1975interpolation}. Hence whenever $A$'s real spectrum lies in $[-\rho(A),\rho(A)]$,
\[
    \rho\bigl(\tilde T_d(A/\rho(A))\bigr) \;\leq\; 2^{1-d},
\]
so the preconditioned dynamics has spectral radius shrunk by a factor of $2^{1-d}$ relative to the original system. Combining this fact with online convex regression delivers the first sublinear, \emph{hidden-dimension-independent} regret bounds for online prediction in the asymmetric LDS class. Empirically, \citet{marsden2024universal} show that USP with fixed Chebyshev or Legendre coefficients reduces prediction error on synthetic linear, nonlinear, and deep-RNN-generated sequences and on the ETTh1 benchmark.

\paragraph{Does this transfer to universal time-series forecasting?}
The setting of pretrained foundation models for time series~\citep{woo2024unified, liu2024moirai2, das2024decoder, ansari2024chronos, goswami2024moment, auer2025tirex} differs from the theoretical framework of USP in three important ways:
\begin{enumerate}[leftmargin=1.5em, itemsep=0.1em, topsep=0.2em]
    \item \emph{The forecaster is a patch-based transformer rather than an online learner.} Models such as Moirai-2~\citep{liu2024moirai2} segment the input into fixed-length patches and embed each patch independently before any attention. The cumulative regret of an online learner is replaced by a batch-style empirical loss, and the model has access to the entire context window simultaneously rather than one observation at a time.
    \item \emph{The data-generating process is far richer than a finite-dimensional LDS.} Pretraining corpora such as LOTSA~\citep{woo2024unified} aggregate billions of observations across heterogeneous domains, frequencies, and noise structures; many of the resulting signals are not well-modelled as outputs of a single low-dimensional LDS, so the algebraic equivalence ``$p(B) y_t \leftrightarrow p(A) x_t$'' applies, at best, only locally and approximately.
    \item \emph{The task is multi-step-ahead forecasting, not single-step online prediction.} A naive approach to USP for forecasting would precondition the target, train the predictor to emit the preconditioned next-step value $\hat{\tilde y}_{t+1}$, and invert the convolution at inference to recover $\hat y_{t+1}$. However, the inversion is a causal recurrence
    \begin{equation}
        \hat y_t \;=\; \hat{\tilde y}_t \;-\; \sum_{k=1}^{d} c_k\, \hat y_{t-k},
        \label{eq:inversion}
    \end{equation}
    so for an autoregressive multi-step rollout the prediction error compounds across the inversion, leading to catastrophic accumulation.
\end{enumerate}

\paragraph{Our approach: a residual preconditioning channel.}
To resolve the multi-step inversion problem, we leave the forecast target unchanged and use the polynomial-preconditioned signal as an \emph{auxiliary input channel}. Concretely, given a univariate input sequence $x_t$, we apply a degree-$d$ monic polynomial $p$ in the lag operator $B_s$ at stride $s$ and form the residual
\begin{equation}
    r_t \;=\; (p(B_s) x)_t - x_t \;=\; \sum_{k=1}^{d} c_k \, x_{t-k s},
    \label{eq:precond-intro}
\end{equation}
where $\{c_k\}_{k=1}^{d}$ are the non-leading coefficients of the chosen monic polynomial and the stride $s$ is set equal to the patch size $P$. We use the residual rather than the full filter output so that the preconditioning channel captures the \emph{novel} information injected by preconditioning (subtracting the raw signal, which is already available as the existing target channel). For our default choice---the monic Chebyshev polynomial $\tilde T_d$ with $d=4$, $s = P = 16$---the residual reduces to
\[
    r_t \;=\; -x_{t-2P} + \tfrac{1}{8} x_{t-4P},
\]
i.e., a sparse FIR filter that injects information from two and four patches in the past. Because the model never operates in preconditioned space at inference, no inversion is needed and the multi-step error accumulation of Eq.~\eqref{eq:inversion} is avoided.

\paragraph{Why this should help: the cross-patch information bottleneck.}
Patch-based architectures~\citep{nie2023patchtst, dosovitskiy2021image} embed each patch independently before attention, creating what we call a \emph{cross-patch information bottleneck}: before the first self-attention layer, a patch embedding encodes only $P$ time steps, so temporal correlations whose natural scale exceeds $P$ (e.g., daily periodicities in $15$-minute data, spanning $96$ steps or $6$ patches at $P=16$) must be discovered end-to-end by attention. A polynomial preconditioner that mixes information across multi-patch scales \emph{before} patching directly alleviates this bottleneck: each patch embedding now sees a low-dimensional summary of the cross-patch context for free. Combined with the spectral-attenuation property of Chebyshev polynomials, this provides a principled, parameter-efficient inductive bias that the model would otherwise have to discover through attention.

\paragraph{Empirical evidence.}
We retrain the Moirai-2 Small encoder~\citep{liu2024moirai2} ($11.4$M parameters, six decoder blocks, $P=16$) from scratch on the LOTSA corpus~\citep{woo2024unified} for $10$K and $100$K optimizer steps, comparing four conditions: a standard \Baseline{}, a capacity-matched \Zero{} control (auxiliary channel filled with zeros), a capacity-matched \Dup{} control (auxiliary channel = copy of input), and our method \HD{} (Chebyshev $d=4$, stride $16$, $10\%$ channel dropout). On the $97$-configuration GIFT-Eval benchmark~\citep{vijay2024gifteval}, at context length $4{,}000$ and with five random seeds, \HD{} achieves $\Delta\mathrm{MASE}^* = -2.9\%$ at $10$K steps and $-3.9\%$ at $100$K steps. The \Dup{} capacity control \emph{hurts} performance ($44.7\%$ win rate) and \Zero{} is neutral ($54\%$), identifying the polynomial content---not the extra parameters---as the source of the gain. Cross-domain analysis is consistent with the spectral hypothesis: gains are largest on high-frequency domains (transport, energy) whose natural time scales align with the filter's lookback structure, and negligible on weather and healthcare where the dominant variations occur at scales either much shorter or much longer than our stride-$16$ filter.

\paragraph{Contributions and outline.}
The remainder of this paper develops the following contributions:
\begin{enumerate}[leftmargin=1.5em, topsep=0.2em, itemsep=0.2em]
  \item A self-contained mathematical treatment (Sections~\ref{sec:preliminaries}--\ref{sec:precondition}) of universal sequence preconditioning specialised from online LDS prediction to patch-based transformer forecasting, including: the Marsden--Hazan algebraic identity, the Chebyshev equioscillation theorem, a closed-form derivation of our filter's frequency response (Theorem~\ref{thm:freq-response}), a BIBO-stability lemma (Lemma~\ref{lem:stability}), a noise-gain summary, and a basis comparison against Legendre.
  \item A self-contained description (Section~\ref{sec:patch-transformers}) of the Moirai-2 architecture, its prefix-LM pretraining objective, the LOTSA training corpus, the multi-token inference procedure, the GIFT-Eval evaluation protocol with a rigorous definition of the MASE metric, and the architectural modification implementing our method (Section~\ref{sec:method}).
  \item An empirical evaluation (Section~\ref{sec:empirical}) on the GIFT-Eval benchmark, in which polynomial input preconditioning improves a capacity-limited patch-based transformer's $\overline{\mathrm{MASE}^*}$ by $\Delta\mathrm{MASE}^* = -2.9\%$, supplemented by extensive capacity-control and filter ablations, an independent FEV-Bench cross-benchmark check, context-length and 100K-step extensions, and per-domain/per-frequency analyses.
\end{enumerate}

%==============================================================================
\section{Preliminaries}
\label{sec:preliminaries}
%==============================================================================

The mathematical background for our method has two roles. First, it explains why long-memory sequence prediction is easy for strictly stable systems but hard near the unit circle. Second, it identifies the specific object we borrow from the theory: a fixed polynomial filter that reshapes the spectrum of a sequence before it is passed to a predictor. We follow the linear-dynamical-system notation of \citet[Chs.~4 and~9]{hazan2025onlinecontrol}: $x_t$ for hidden state, $u_t$ for control, $y_t$ for observation, and $w_t, v_t$ for process and observation noise. The patch-based transformer architecture and the specific Moirai-2 instantiation we use are deferred to Section~\ref{sec:patch-transformers}. We present only the statements needed for the paper; exact constants and full proofs are deferred to the cited sources, and a reader familiar with any of these may skim the corresponding subsection.

\subsection{Passive Prediction in Linear Dynamical Systems}
\label{sec:passive-lds}

A partially observed linear time-invariant system has hidden state $x_t \in \R^{d_x}$, observed input $u_t \in \R^{d_u}$, and output $y_t \in \R^{d_y}$:
\begin{equation}
    x_{t+1} = A x_t + B u_t + w_t,
    \qquad y_t = C x_t + v_t,
    \qquad t \in \Z_{\geq 0},
    \label{eq:lds}
\end{equation}
with initial state $x_0 \in \R^{d_x}$. Here $A, B, C$ are unknown system matrices, while $w_t$ and $v_t$ denote process and observation noise. In time-series forecasting there may be no exogenous input; this is recovered by setting $u_t \equiv 0$ or by treating past observations as the relevant input history.

The prediction problem is \emph{passive}: unlike control, the learner does not choose $u_t$ to influence the state. At each time $t$, the learner observes the past history and predicts $\hat y_t$; after observing $y_t$, it suffers loss $\ell_t(\hat y_t, y_t)$, usually squared error. For a predictor class $\Pi$, regret is
\begin{equation}
    \mathrm{Regret}_T(\mathcal{A}, \Pi) \;:=\; \sum_{t=1}^{T} \ell_t(\hat y_t, y_t) - \min_{\pi \in \Pi} \sum_{t=1}^{T} \ell_t(\hat y_t^{\pi}, y_t).
    \label{eq:prediction-regret}
\end{equation}
Sublinear regret means that the learner's average loss approaches that of the best predictor in the reference class. This is the sense in which the online-learning results below should be read.

\subsection{Stability and Finite Memory}
\label{sec:stability-memory}

The \emph{spectral radius} of a square matrix is $\rho(A) := \max\{\abs{\lambda} : \lambda \in \mathrm{spec}(A)\}$. For a time-invariant linear system, $\rho(A) < 1$ is equivalent to $A^t \to 0$ as $t \to \infty$. For finite-time learning bounds one usually needs the quantitative version
\begin{equation}
    \norm{A^t} \;\leq\; M \rho^t \qquad \text{for all } t \geq 0, \qquad \rho \in (0, 1),
    \label{eq:quantitative-stability}
\end{equation}
which says that the influence of the remote past decays geometrically. The following truncation lemma is the precise version of the common statement that strictly stable systems have finite effective memory.

\begin{lemma}[Stable LDSs are approximated by finite-memory predictors; cf.\ {\citealp[Lemma~9.6]{hazan2025onlinecontrol}}]
\label{lem:stable-truncation}
Consider the noiseless input-output LDS $x_{t+1} = A x_t + B u_t$, $y_t = C x_t$. Suppose $\norm{A^t} \leq M \rho^t$, $\norm{B} \leq \beta$, $\norm{C} \leq \gamma$, $\norm{u_t} \leq U$ with $\rho \in (0, 1)$. Let $\pi_{A,B,C,x_0}$ denote the exact open-loop LDS predictor and let $\pi_{A,B,C,h}$ be the length-$h$ truncated predictor $\hat y_t^{(h)} = \sum_{i=1}^{h} CA^{i-1}B u_{t-i}$. Then
\begin{equation}
    \sum_{t=1}^{T} \norm{\hat y_t^{\pi_{A,B,C,x_0}} - \hat y_t^{(h)}} \;\leq\; T \cdot \frac{M \beta \gamma U}{1 - \rho} \rho^h + \sum_{t=1}^{\min\{T, d_x\}} \norm{C A^t x_0}.
    \label{eq:stable-truncation}
\end{equation}
\end{lemma}

Thus, after the initial transient, a stable system can be approximated to error $\varepsilon$ using memory $h = O(\log(T/\varepsilon)/(1 - \rho))$ when $\rho$ is close to one. This is the useful ``strictly stable'' baseline: finite-memory linear prediction is valid, but the required memory can grow poorly as the stability margin $1 - \rho$ shrinks.

When $\rho(A) = 1$ (marginally stable, e.g., pure sinusoids or integrated processes), this geometric truncation no longer gives a finite-memory approximation with rapidly vanishing tail; classical regret bounds degrade, with the dependence on $d_x$ often becoming linear or worse. The methods we develop are designed for exactly this difficult regime.

\subsection{Linear Predictors and Online Convex Optimization}
\label{sec:linear-predictors}

The standard convex comparator class for passive prediction is the class of finite-memory linear predictors:
\begin{definition}[Finite-memory linear predictors]
\label{def:linear-predictors}
For memory lengths $h, k$ and norm radius $\kappa$, define $\Pi^L_{h, k, \kappa}$ to be the set of predictors
\begin{equation}
    \hat y_t \;=\; \sum_{i=1}^{h} M_{1, i} \, y_{t-i} + \sum_{j=1}^{k} M_{2, j} \, u_{t-j},
    \qquad
    \sum_{i=1}^{h} \norm{M_{1, i}} + \sum_{j=1}^{k} \norm{M_{2, j}} \leq \kappa.
    \label{eq:linear-predictor}
\end{equation}
\end{definition}

This class is computationally attractive because it is convex in the predictor matrices. Online Gradient Descent (OGD) therefore gives an immediate learning guarantee.

\begin{theorem}[OGD for finite-memory predictors; cf.\ {\citealp[Thm.~6.1 and Cor.~9.7]{hazan2025onlinecontrol}}]
\label{thm:ogd-linear-predictors}
Let $K$ be a convex parameter set of diameter $D$, and suppose the loss gradients over $K$ have norm at most $G$. OGD with step sizes $\eta_t = D / (G \sqrt{t})$ satisfies
\begin{equation}
    \sum_{t=1}^{T} \ell_t(M_t) - \min_{M \in K} \sum_{t=1}^{T} \ell_t(M) \;\leq\; \frac{3}{2} G D \sqrt{T}.
    \label{eq:ogd-regret}
\end{equation}
In particular, finite-memory linear predictors are learnable with $O(\sqrt{T})$ regret, with constants depending on the memory length, norm radius, and data bounds.
\end{theorem}

There is also an exact structural reason to study linear predictors. In the noiseless case, every finite-dimensional LDS induces a finite-memory autoregressive predictor.

\begin{theorem}[Cayley--Hamilton representation; cf.\ {\citealp[Thm.~9.4]{hazan2025onlinecontrol}}]
\label{thm:cayley-hamilton}
Consider the noiseless LDS $x_{t+1} = A x_t + B u_t$, $y_t = C x_t$, where $A \in \R^{d_x \times d_x}$ has characteristic polynomial $p_A(\lambda) = \lambda^{d_x} + a_1 \lambda^{d_x - 1} + \cdots + a_{d_x}$. Then there exist matrices $M_{1, i} \in \R^{d_y \times d_y}$ and $M_{2, j} \in \R^{d_y \times d_u}$, $i, j = 1, \ldots, d_x$, such that for every $t > d_x$,
\begin{equation}
    y_t \;=\; \sum_{i=1}^{d_x} M_{1, i} \, y_{t-i} + \sum_{j=1}^{d_x} M_{2, j} \, u_{t-j}.
    \label{eq:cayley-hamilton-predictor}
\end{equation}
Concretely, one may take $M_{1, i} = -a_i I_{d_y}$ and $M_{2, j} = C\bigl(\sum_{i=0}^{j-1} a_i A^{j-1-i}\bigr) B$ (with $a_0 = 1$).
\end{theorem}

Theorem~\ref{thm:cayley-hamilton} says that linear predictors are not merely a heuristic: in the noiseless setting they exactly represent finite-dimensional LDS predictions after a transient, with autoregressive coefficients given precisely by the coefficients of $A$'s characteristic polynomial. This identity is the algebraic foundation behind why convolving the observed sequence with a polynomial---preconditioning---is meaningful: it operates on the same finite-memory representation that is, by Cayley--Hamilton, exact. The drawback is that the memory length depends on the hidden dimension $d_x$. Lemma~\ref{lem:stable-truncation} removes this dependence for strictly stable systems, but replaces it with dependence on the stability margin $1 - \rho$. \emph{Spectral filtering} addresses precisely this bottleneck.

\subsection{Spectral Filtering: Compressing Long Memory}
\label{sec:spectral-filtering-prelim}

Spectral filtering was introduced by \citet{hazan2017learning} to learn LDSs without explicit system identification and without a long finite-memory window; it is the immediate mathematical predecessor of universal sequence preconditioning. The exposition in \citet[Ch.~11]{hazan2025onlinecontrol} begins with the scalar impulse-response vectors
\begin{equation}
    \mu_\alpha \;=\; \begin{bmatrix} 1 & \alpha & \alpha^2 & \cdots & \alpha^{T-1} \end{bmatrix}^\top, \qquad \alpha \in [0, 1],
    \label{eq:impulse-vector}
\end{equation}
and the Hankel/Hilbert matrix
\begin{equation}
    Z_T \;:=\; \int_0^1 \mu_\alpha \mu_\alpha^\top \, d\alpha \in \R^{T \times T},
    \qquad (Z_T)_{ij} = \frac{1}{i + j + 1} \quad (i, j = 0, \ldots, T - 1).
    \label{eq:hankel-matrix}
\end{equation}
Let $\phi_1, \ldots, \phi_T$ be the eigenvectors of $Z_T$, ordered by decreasing eigenvalue. The eigenvalues of $Z_T$ decay rapidly:
\begin{equation}
    \sigma_k(Z_T) \;\leq\; 4 \pi^2 \exp\!\left(-\frac{k}{\log T}\right) \qquad (k, T > 1),
    \label{eq:hankel-decay}
\end{equation}
as stated in \citet[Thm.~11.1]{hazan2025onlinecontrol}. This decay implies that the first polylogarithmically many eigenvectors approximate the family of impulse responses $\{\mu_\alpha : \alpha \in [0, 1]\}$ well.

The resulting spectral predictor uses the features $\langle \tilde u_t, \phi_i \rangle$, where $\tilde u_t = (u_{t-1}, u_{t-2}, \ldots, u_1, 0, \ldots, 0)$. In the scalar case,
\begin{equation}
    \hat y_t \;=\; \sum_{i=1}^{h} M_i \, \langle \tilde u_t, \phi_i \rangle.
    \label{eq:spectral-predictor}
\end{equation}
For symmetric LDSs with real spectrum, diagonalising $A$ reduces the high-dimensional impulse response to a sum of scalar responses of the form~\eqref{eq:impulse-vector}, and the parameter count $h = O(\log^2 T / \varepsilon)$ suffices to $\varepsilon$-approximate the entire optimal-predictor class \emph{independent of} $d_x$ and $1 - \rho$. The online learner then runs a convex online algorithm over the coefficients $M_i$.

\begin{remark}[What spectral filtering contributes]
Spectral filtering replaces a long raw-history basis by a compressed basis tailored to LDS impulse responses. It is the immediate mathematical predecessor of the present work. Our method does not learn Hankel eigenfeatures; instead, it injects one fixed polynomially filtered channel into a transformer. The shared idea is that a carefully chosen linear transformation of the input sequence can expose long-range dynamical structure to a downstream learner.
\end{remark}

The preceding stable-system picture becomes significantly less favourable when eigenvalues approach the unit circle, and even more so when $A$ is asymmetric. \citet[Thm.~12.1]{hazan2025onlinecontrol} gives a cyclic-permutation LDS construction in which any online learner incurs expected cumulative squared loss $\Omega(d_x)$ over $T \geq d_x$ rounds before the hidden state is revealed. The asymmetric extension of spectral filtering is more delicate; the universal-sequence-preconditioning technique we describe in Section~\ref{sec:precondition} is one approach to this regime, and is the immediate mathematical foundation of our method.

\subsection{Chebyshev Polynomials}
\label{sec:chebyshev-prelim}

The \emph{Chebyshev polynomials of the first kind}, denoted $T_d(z)$ for $d \in \Z_{\geq 0}$, are defined by the recurrence
\begin{equation}
    T_0(z) = 1, \quad T_1(z) = z, \quad T_{d+1}(z) = 2 z \, T_d(z) - T_{d-1}(z),
    \label{eq:cheb-recurrence}
\end{equation}
or equivalently by the trigonometric identity $T_d(\cos \theta) = \cos(d \theta)$. The first several polynomials are $T_2(z) = 2z^2 - 1$, $T_3(z) = 4z^3 - 3z$, $T_4(z) = 8z^4 - 8z^2 + 1$. The leading coefficient of $T_d$ is $2^{d-1}$, so the \emph{monic Chebyshev polynomial} is $\tilde T_d(z) = T_d(z) / 2^{d-1}$. For $d = 4$,
\[
    \tilde T_4(z) = z^4 - z^2 + \tfrac{1}{8}.
\]

Chebyshev polynomials possess several extremal properties; one of the most celebrated is the \emph{equioscillation property} underlying the approximation theorem we state in Section~\ref{sec:chebyshev}.

%==============================================================================
\section{Universal Sequence Preconditioning}
\label{sec:precondition}
\label{sec:chebyshev}
%==============================================================================

\citet{marsden2024universal} introduced a polynomial-preconditioning framework that yields the first sublinear, hidden-dimension-independent regret bounds for online prediction in marginally-stable LDSs. The mechanism: convolving the observation sequence with a fixed degree-$d$ monic polynomial $p$ is algebraically equivalent to multiplying the hidden dynamics by $p^*(A)$, where $p^*(z) := z^d p(1/z)$ is the reciprocal polynomial; choosing $p$ to be the monic Chebyshev polynomial then minimax-optimally compresses the spectral radius the downstream learner faces. We extract from this framework the analytic ingredients needed for our patch-based forecasting setting: the algebraic identity (Section~\ref{sec:lds-identity}), the Chebyshev minimax theorem and its corollary on spectral compression (Section~\ref{sec:cheb-minimax}), the frequency response and stability properties of the resulting FIR filter (Section~\ref{sec:freq-response-section}), and the comparison with other polynomial bases (Section~\ref{sec:basis-comparison}). The full Marsden--Hazan regret machinery---a careful summation of the spectral residual under marginal stability---is not used in our analysis; the reader interested in those rates is referred to \citet[Thms.~3 and~5]{marsden2024universal}.

\subsection{Polynomial Preconditioning of LDS Observations}
\label{sec:lds-identity}

Consider an LDS as in~\eqref{eq:lds} with noise terms suppressed for exposition, so that $y_t = C A^t x_0$. For a monic polynomial $p(z) = \sum_{k=0}^{d} a_k z^k$ with $a_0 = 1$ and $a_d = 1$, the preconditioned observation is
\begin{equation}
    \tilde y_t \;:=\; (p(B) y)_t \;=\; \sum_{k=0}^{d} a_k\, y_{t-k} \;=\; C \cdot p^*(A) \cdot A^{t-d} x_0,
    \label{eq:mh-identity}
\end{equation}
where $B$ is the backshift operator and $p^*(z) := z^d p(1/z)$. The identity replaces the implicit dependence of $\tilde y_t$ on $A^{t-d}$ by $p^*(A) \cdot A^{t-d}$. If $A$'s real spectrum lies in $[-\rho, \rho]$, the relevant bound is
\begin{equation}
    \rho\big( p^*(A) \big) \;\leq\; \sup_{\lambda \in [-\rho,\rho]} \abs{p(\lambda)},
    \label{eq:rho-bound}
\end{equation}
since $p$ and $p^*$ have the same coefficient set up to reversal. Minimising the right-hand side over monic $p$ of degree $d$ is the classical Chebyshev minimax problem (Theorem~\ref{thm:chebyshev-minimax}), and its solution $p(\lambda) = \rho^d \tilde T_d(\lambda/\rho)$ achieves $\sup_{[-\rho,\rho]} \abs{p} \leq \rho^d \cdot 2^{1-d}$. Preconditioning by Chebyshev coefficients therefore compresses the effective spectral radius seen by a downstream predictor by a factor of $2^{1-d}$ in the worst case over $A \in \{A : \mathrm{spec}(A) \subset [-\rho,\rho]\}$.

\paragraph{Specialization to patch-based forecasting.}
Our setting differs from the LDS online-prediction setting in three ways: (i) the forecaster is a pretrained transformer, not an online learner; (ii) the data-generating process is unknown and not assumed to be a finite-dimensional LDS; (iii) rather than \emph{replacing} the target sequence by its preconditioned version (which would require inverting the convolution at inference and is incompatible with multi-step rollout), we \emph{concatenate} the preconditioned sequence as an auxiliary input channel. Concretely, we apply the preconditioning operator at stride $s$ to the input series $x$ and feed the model the residual
\begin{equation}
    r_t \;=\; (p(B_s) x)_t - x_t \;=\; \sum_{k=1}^{d} c_k\, x_{t-ks},
    \label{eq:residual-channel}
\end{equation}
alongside the original $x$. We work with the residual rather than the full filter output because the original signal is already provided by the unmodified target channel; the residual captures the \emph{novel} cross-patch information injected by preconditioning. Despite the differences from the original LDS setting, the spectral intuition transfers: a Chebyshev preconditioner shapes the input's spectrum minimax-optimally, and a patch-based transformer facing a signal whose spectrum has been shaped to highlight multi-patch correlations is plausibly better-equipped to capture them.

\subsection{Chebyshev Minimax}
\label{sec:cheb-minimax}

\begin{theorem}[Chebyshev minimax; cf.\ \citealp{davis1975interpolation}]
\label{thm:chebyshev-minimax}
Among all real monic polynomials of degree $d \geq 1$, the monic Chebyshev polynomial $\tilde T_d$ uniquely minimises $\sup_{x \in [-1,1]} \abs{q(x)}$:
\[
    \tilde T_d \;=\; \argmin_{q \text{ monic, } \deg q = d} \; \max_{x \in [-1,1]} \abs{q(x)}, \qquad \max_{x \in [-1,1]} \abs{\tilde T_d(x)} = 2^{1-d}.
\]
\end{theorem}

\begin{proof}[Proof sketch]
By the trigonometric definition $T_d(\cos\theta) = \cos(d\theta)$, $T_d$ ranges over $[-1,1]$ on $[-1,1]$ with $d+1$ alternating extrema $\tilde T_d(x_k) = (-1)^k \cdot 2^{1-d}$ at $x_k = \cos(k\pi/d)$, $k = 0, \ldots, d$. Suppose for contradiction that some monic $q$ of degree $d$ has $\sup_{[-1,1]} \abs{q} < 2^{1-d}$. Then $r := \tilde T_d - q$, of degree $\leq d-1$, takes the same sign as $\tilde T_d$ at each $x_k$ and so alternates in sign across $d+1$ points; the intermediate-value theorem gives at least $d$ zeros, forcing $r \equiv 0$, a contradiction.
\end{proof}

\begin{corollary}[Worst-case spectral compression]
\label{cor:spectral-compression}
For diagonalisable $A$ with real spectrum in $[-\rho, \rho]$, the choice $q(\lambda) = \rho^d \tilde T_d(\lambda/\rho)$ minimises $\sup_A \rho(q(A))$ over monic polynomials of degree $d$, attaining
\[
    \sup \rho\big(q(A)\big) \;=\; \sup_{\lambda \in [-\rho, \rho]} \abs{q(\lambda)} \;\leq\; \rho^d \cdot 2^{1-d}.
\]
\end{corollary}

\subsection{Frequency Response, Stability, and Residual Filter}
\label{sec:freq-response-section}

For our patch-based setting, the filter is applied at stride $s = P$ and we use the reciprocal monic Chebyshev $p(z) := \tilde T_d^*(z) = z^d \tilde T_d(1/z)$. For $d=4$ explicitly: $p(z) = 1 - z^2 + z^4/8$. Let $H(\omega) := p(e^{-j 2\pi s \omega})$ denote the DTFT of the stride-$s$ filter.

\begin{theorem}[Frequency response]
\label{thm:freq-response}
Let $p = \tilde T_d^*$ and $H(\omega) = p(e^{-j 2\pi s \omega})$. Then:
\begin{enumerate}[leftmargin=1.5em, itemsep=0.1em]
    \item $\abs{H(\omega)}^2$ is a real trigonometric polynomial in $\cos(2\pi s\omega)$ of degree $d$, periodic in $\omega$ with period $1/s$.
    \item $H$ has no zeros on the unit circle: the real roots of $\tilde T_d$ lie in $(-1,1)$, so the roots of $\tilde T_d^*$ lie outside $\{\abs{z}=1\}$.
    \item For even $d$, at the harmonics $\omega_k := k/(2s)$, $k \in \Z$, the magnitude attains the Chebyshev minimax constant
    \begin{equation}
        \abs{H(\omega_k)} \;=\; \abs{p(\pm 1)} \;=\; 2^{\,1-d},
        \label{eq:min-magnitude}
    \end{equation}
    and these are critical points of $\abs{H(\omega)}^2$.
\end{enumerate}
\end{theorem}

\begin{proof}
Part (i) follows by expanding $\abs{H}^2 = H\bar H$ and grouping conjugate pairs $a_k a_\ell e^{j(k-\ell)\phi}$, $\phi = 2\pi s\omega$, then applying the Chebyshev identity $\cos(n\phi) = T_n(\cos\phi)$. Part (ii) is the algebraic fact that reciprocation $z \mapsto 1/z$ sends real roots in $(-1,1)\setminus\{0\}$ to reals with $\abs{z} > 1$. Part (iii): for even $d$, odd-degree coefficients of $\tilde T_d$ vanish, so $p(\pm 1) = T_d(\pm 1)/2^{d-1} = 2^{1-d}$, and at $\omega_k$ we have $e^{-j 2\pi s \omega_k} = (-1)^k$.
\end{proof}

For our default setting $d = 4$, $s = 16$: the attenuation harmonics lie at $\omega \in \{0, 1/32, 1/16, 3/32, \ldots\}$ with $\abs{H(\omega_k)} = 1/8$. The filter's two non-trivial taps are at lags $2s = 32$ and $4s = 64$ time steps---two and four patches in the past for $P = 16$.

\begin{lemma}[BIBO stability and variance bound]
\label{lem:stability}
Let $h_t = (\tilde T_d^*(B_s) x)_t$ for an input $x$ with $\sup_t \abs{x_t} \leq M$ and $\Var(x_t) \leq \sigma^2$. Then
\[
    \sup_t \abs{h_t} \;\leq\; C_d M, \qquad \Var(h_t) \;\leq\; C_d^2 \sigma^2, \qquad C_d := \sum_{k=0}^{d} \abs{a_k}.
\]
For $d = 4$, $C_4 = 1 + 1 + 1/8 = 2.125$, so $\sup_t \abs{h_t} \leq 2.125 M$ and $\Var(h_t) \leq 4.52 \sigma^2$.
\end{lemma}

\noindent The lemma is immediate from the triangle inequality applied to $h_t = \sum_k a_k x_{t-ks}$ and ensures that the preconditioning channel does not cause pathological numerical behaviour.

\paragraph{The residual filter.}
The model receives the residual $r_t = ((p - 1)(B_s) x)_t$, not the raw filter output. Its DTFT is $\widetilde H(\omega) = H(\omega) - 1$, a band-pass-like operator that passes (gain $\geq 1$) frequencies where $H$ has gain $\geq 2$ and attenuates frequencies that $H$ attenuates. Specifically:

\begin{proposition}[Residual filter at the attenuation harmonics]
\label{prop:residual-filter}
At the harmonics $\omega_k = k/(2s)$ of Theorem~\ref{thm:freq-response},
\[
    \abs{\widetilde H(\omega_k)} \;=\; \abs{H(\omega_k) - 1} \;=\; \abs{2^{1-d} - 1}.
\]
For $d = 4$, $\abs{\widetilde H(\omega_k)} = 7/8$.
\end{proposition}

\subsection{Noise Gain and Comparison with Other Bases}
\label{sec:basis-comparison}
\label{sec:noise-gain}

The spectral compression of Theorem~\ref{thm:chebyshev-minimax} comes at a cost: the filter amplifies independent noise by a factor proportional to $\norm{\boldsymbol{a}^{(d)}}_2^2$, the squared $\ell^2$ norm of the monic Chebyshev coefficients in the monomial basis. The first few values are
\[
    \norm{\boldsymbol{a}^{(d)}}_2^2 \;=\; 1,\; 1,\; \tfrac{5}{4},\; \tfrac{25}{16},\; \tfrac{129}{64},\; \tfrac{681}{256},\; \tfrac{3653}{1024},\; \tfrac{19825}{4096},\; \tfrac{108545}{16384}, \qquad d = 0, \ldots, 8,
\]
with consecutive ratios approaching $((1+\sqrt 2)/2)^2 \approx 1.457$. The noise-gain norm grows exponentially in $d$ at rate strictly slower than the spectral-compression rate $4^{1-d}$ shrinks, so the spectral benefit dominates the noise penalty for moderate $d$. At $d = 4$: spectral compression $4^{-3} = 1/64$ versus noise gain $\norm{\boldsymbol{a}^{(4)}}_2^2 = 129/64 \approx 2.02$. Empirically (Table~\ref{tab:ablations-main}(a)), the MASE curve over $d \in \{2, \ldots, 8\}$ is shallow with $d = 4$ achieving the lowest mean, consistent with the spectral term being deeply saturated by $d \geq 3$ on capacity-limited patch transformers.

\paragraph{Comparison with Legendre.}
The minimax property singles out Chebyshev among classical orthogonal polynomial families. Different orthogonality weights yield different optimisation criteria: Legendre $\tilde P_d$ minimises the $L^2$ norm $(\int_{-1}^1 q^2)^{1/2}$ rather than $L^\infty$. Since $\rho(p^*(A)) \leq \sup_{[-1,1]} \abs{p}$ is an $L^\infty$ bound, Chebyshev gives the right inductive bias for diagonalisable $A$ with bounded spectrum.

\begin{proposition}[Sup-norm gap between Chebyshev and Legendre]
\label{prop:cheb-vs-legendre}
For $d \geq 1$,
\[
    \frac{\sup_{[-1,1]} \abs{\tilde T_d}}{\sup_{[-1,1]} \abs{\tilde P_d}} \;=\; \frac{2 \binom{2d}{d}}{4^d} \;\xrightarrow[d \to \infty]{}\; \frac{2}{\sqrt{\pi d}}.
\]
The ratio equals $1$ at $d = 1$ and is strictly less than $1$ for all $d \geq 2$, with values $1, 3/4, 5/8, 35/64, 63/128, \ldots$ for $d = 1, 2, 3, 4, 5, \ldots$.
\end{proposition}

\begin{proof}
Theorem~\ref{thm:chebyshev-minimax} gives $\sup_{[-1,1]} \abs{\tilde T_d} = 2^{1-d}$. For Legendre: $\abs{P_d(x)} \leq 1$ on $[-1,1]$ with equality at $x = \pm 1$, so the monic $\tilde P_d = P_d / c_d$ with leading coefficient $c_d = (2d)!/(2^d (d!)^2)$ has sup norm $1/c_d = 2^d (d!)^2 / (2d)!$. Taking the ratio and applying Stirling gives the stated asymptotic.
\end{proof}

For $d = 4$, the monic Chebyshev attenuates a worst-case eigenvalue by $1/8 = 0.125$, while the monic Legendre attenuates by only $8/35 \approx 0.229$---a $1.83\times$ gap. Squared, this is roughly a $3.35\times$ gap in the spectral-residual term.

\paragraph{Empirical confirmation.} The basis ablation (Appendix~\ref{app:ablations}, Table~\ref{tab:basis-ablation}) is consistent:
\begin{center}
\small
\begin{tabular}{lcc}
\toprule
Basis ($d{=}4$, $s{=}16$) & MASE & $\Delta\%$ vs.\ Baseline \\
\midrule
Chebyshev $\tilde T_4^*$ & $0.8234$ & $-4.99\%$ \\
EMA (truncated geometric) & $0.8329$ & $-3.89\%$ \\
Legendre $\tilde P_4^*$ & $0.8400$ & $-3.07\%$ \\
Finite differences & $0.8569$ & $-1.12\%$ \\
\bottomrule
\end{tabular}
\end{center}

The ordering Chebyshev $>$ EMA $>$ Legendre $>$ finite differences confirms the theoretical prediction: the family that minimises the $L^\infty$ norm on $[-1,1]$ outperforms families optimised for other norms.

\paragraph{Why not learn the coefficients?} A natural alternative is to drop the fixed Chebyshev structure and let the model learn the FIR coefficients from initialisation. Empirically (Appendix~\ref{app:ablations}, Table~\ref{tab:learned-fir}), learnable coefficients converge to a high-pass-like solution that beats baseline by $-3.85\%$ but is worse than fixed Chebyshev ($-4.99\%$). Fixing the polynomial provides an inductive bias the model would otherwise have to discover from data---consistent with the broader theme that classical analytic structure (orthogonal polynomials, equioscillation) is more sample-efficient than free parameter optimisation when the structure is correct.

%==============================================================================
\section{Patch-based Transformers for Forecasting}
\label{sec:patch-transformers}
%==============================================================================

Our experiments are conducted on the \emph{Moirai-2 Small} time series foundation model~\citep{liu2024moirai2}, the open-source\footnote{Available on the HuggingFace Hub as \texttt{Salesforce/moirai-2.0-R-small}; the architecture and training recipe follow the description in \citet{liu2024moirai2}.} dense, decoder-only successor to the original Moirai encoder~\citep{woo2024unified}. We chose Moirai-2 because it is a small ($11.4$M parameter), publicly reproducible, state-of-the-art universal forecaster: small enough to train from scratch on a single H100 within wall-clock budgets compatible with five-seed ablations, yet competitive with much larger models on the GIFT-Eval benchmark. Our preconditioning method modifies only the input projection of this model; understanding the surrounding architecture, training objective, training corpus, and evaluation protocol is therefore essential context for the empirical results in Section~\ref{sec:empirical}. We summarise each in turn.

\subsection{Patches and the Per-Patch Projection}
\label{sec:patches}

A patch-based transformer forecaster~\citep{nie2023patchtst, woo2024unified, liu2024moirai2} proceeds as follows. Given a univariate history $x_{1:T}$, partition it into non-overlapping patches $\mathbf{p}_i = (x_{(i-1)P+1}, \ldots, x_{iP})$ of length $P$ for $i=1, \ldots, T/P$. A per-patch projection (in Moirai-2, a residual block of two linear layers with a SiLU nonlinearity) maps each patch to an embedding $\mathbf{e}_i \in \R^{d_{\text{model}}}$. These embeddings are fed to a transformer block stack whose output is decoded into a predictive distribution over future patches via a per-patch projection head. Two features of this architecture are salient for our development:
\begin{itemize}[leftmargin=1.5em, itemsep=0.1em, topsep=0.2em]
    \item \emph{Cross-patch information must be learned via attention.} Before the first self-attention operation, $\mathbf{e}_i$ is a function of $\mathbf{p}_i$ alone; it contains no information about $\mathbf{p}_{i-1}$, $\mathbf{p}_{i+1}$, or more distant patches. Patterns whose natural temporal scale exceeds $P$ time steps (e.g., daily periodicities in $15$-minute data, spanning $96$ steps or $6$ patches at $P=16$) must be discovered end-to-end by the attention layers above the projection.
    \item \emph{The per-patch projection is narrow.} For Moirai-2 Small ($P=16$, $d_{\text{model}}=384$), the projection's input dimension is $2P = 32$ (target plus observation mask), so it has roughly $32 \cdot 384 \approx 12{,}000$ parameters out of $11.4$M total ($\sim 0.1\%$). Adding a third $P$-dimensional input channel costs an additional $P \cdot d_{\text{model}} = 6{,}144$ parameters; altering downstream layers, in contrast, is much more expensive.
\end{itemize}
These two observations motivate \emph{input-side} preconditioning: a fixed FIR filter applied before patching can inject cross-patch information into each $\mathbf{e}_i$ at negligible parameter cost, leaving the rest of the network untouched.

\subsection{The Moirai-2 Small Architecture}
\label{sec:moirai-arch}

Moirai-2 Small is a decoder-only transformer with $L = 6$ layers, hidden dimension $d_{\text{model}} = 384$, feed-forward dimension $d_{\text{ff}} = 1024$, $h = 6$ attention heads, and patch size $P = 16$. Each transformer block applies pre-norm self-attention with rotary position embeddings (RoPE)~\citep{su2024rope} and a SwiGLU feed-forward network~\citep{shazeer2020glu}, with RMSNorm~\citep{zhang2019rms} replacing LayerNorm throughout. A binary attention bias~\citep{liu2024moirai2} discriminates between positions in the observed prefix and positions in the prediction horizon. Inputs are $z$-scored at the sequence level before patching using a packed-sequence variant of standard $z$-score scaling. Three features of the architecture are particularly relevant to the rest of this paper:
\begin{itemize}[leftmargin=1.5em, itemsep=0.1em, topsep=0.2em]
    \item \emph{Decoder-only with causal mask.} Unlike the masked-language-model encoder used in Moirai 1.0~\citep{woo2024unified}, Moirai-2 attends to past patches only. Predictions for patch $i$ depend on $\mathbf{e}_1, \ldots, \mathbf{e}_i$ but not on later patches.
    \item \emph{Multi-token prediction head.} The output projection emits, for each input patch, predictions for the next $K = 4$ patches at $9$ quantile levels $\{0.1, 0.2, \ldots, 0.9\}$, yielding $K \cdot 9 \cdot P = 576$ output values per patch.
    \item \emph{Per-patch input residual block.} The input projection is a two-layer residual MLP $\mathrm{ResBlock}_{2P \to d_{\text{model}}}$ acting on the concatenation $[\mathbf{p}_i; \mathbf{m}_i]$ of patch and observation mask. This is the only layer that touches raw patch values; it is also the only layer modified by our method (Section~\ref{sec:method} below).
\end{itemize}

\subsection{Pretraining Objective and Optimization}
\label{sec:pretraining-objective}

Each training sample is a packed sequence of patched univariate series. A random \emph{prefix ratio} (default $0.3$) determines how much of the sequence is treated as observed context; the remainder is the prediction window. Within the prediction window, the target patches are replaced by a learnable mask token, and the model is trained to predict their values. Additionally, with probability $\texttt{mask\_ratio} = 0.5$, individual patches inside the prediction window are corrupted to encourage robustness to missing observations.

The training loss is the \emph{packed quantile MAE loss},
\begin{equation}
    \mathcal{L} \;=\; \frac{1}{N \cdot |\mathcal{Q}|} \sum_{n,t,q} \rho_q\!\left(\, y_{n,t} - \hat y_{n,t}^{(q)} \,\right),
    \label{eq:quantile-loss}
\end{equation}
where the sum runs over patches in the prediction window of all packed samples, $\mathcal{Q} = \{0.1, \ldots, 0.9\}$ is the set of quantile levels, $\hat y_{n,t}^{(q)}$ is the model's $q$-quantile prediction, and $\rho_q(u) = u(q - \mathbf{1}[u < 0])$ is the standard pinball loss~\citep{koenker1978regression}. We use AdamW~\citep{loshchilov2019adamw} with $\beta_1 = 0.9$, $\beta_2 = 0.98$, $\epsilon = 10^{-6}$, weight decay $0.1$, peak learning rate $\eta = 10^{-3}$, $1{,}000$-step linear warmup, and cosine annealing to zero. Training is performed in bf16 mixed precision with batch size $256$ on a single H100 GPU. Anomaly filtering~\citep{liu2024moirai2} resamples training windows whose normalised values exceed a $z$-score threshold of $8.0$ to suppress extreme outliers in LOTSA. The full set of hyperparameters appears in Appendix~\ref{app:methodology}.

\subsection{Training Corpus: LOTSA}
\label{sec:lotsa}

Moirai-2 is pretrained on the \emph{Large-scale Open Time Series Archive} (LOTSA), introduced by \citet{woo2024unified}, which aggregates $\sim 27$\,billion observations across $9$ broad domains: energy, transport, climate, cloud operations, web, sales, healthcare, nature, and economic/financial. Domains are unbalanced---a few datasets (e.g., Buildings-900K, ERA5) dominate raw observation counts---so Moirai-2 uses \emph{variate-proportional weighted sampling}: for each dataset, the sampling weight is inversely proportional to its raw observation count and proportional to the number of distinct variates, which roughly equalises the contribution of large multivariate datasets and small univariate ones. We use the released LOTSA v1 weighting scheme without modification (the \texttt{lotsa\_v1\_moirai2} configuration in the official codebase). Throughout training, each sample is normalised by its own mean and standard deviation (computed within the observed prefix) before patching.

\subsection{Inference and Forecasting}
\label{sec:inference}

At inference, the model is presented with the full observed history followed by mask tokens for the prediction window. Because the multi-token head emits a $K{=}4$-patch forecast for every input patch, predictions for the prediction-window patches are read directly from the corresponding output positions; no autoregressive rollout is required for horizons of up to $K \cdot P = 64$ time steps. For longer horizons, the predicted patches are appended to the context and the procedure repeats. Each forecast at horizon $h$ produces a vector of $9$ quantile estimates at each time step; for our headline metric (MASE), we use the median ($q = 0.5$) as the point forecast.

\subsection{Evaluation: GIFT-Eval}
\label{sec:gifteval}

We evaluate on \textsc{GIFT-Eval}~\citep{vijay2024gifteval}, an open benchmark of $97$ dataset $\times$ horizon configurations spanning $24$ datasets and $7$ frequency classes (sub-hourly through yearly), curated to test the zero-shot generalization claim of universal forecasters. Each configuration specifies a held-out split, prediction horizon, and patch-size convention; we follow the recommended convention of $P = 8$ for quarterly data and $P = 32$ for all other frequencies, with context length $4{,}000$. The remainder of this subsection rigorously defines the metric we report.

\paragraph{Mean Absolute Scaled Error (MASE).}
For a single test series with point forecast $\hat y_{t}$ and ground-truth $y_{t}$ over a prediction window of length $H$, the \emph{Mean Absolute Scaled Error} \citep{hyndman2006another} is defined as
\begin{equation}
    \mathrm{MASE}(\hat y, y) \;=\; \frac{\frac{1}{H}\sum_{t=1}^{H} \abs{y_t - \hat y_t}}{\frac{1}{N - m}\sum_{t = m+1}^{N} \abs{y^{\text{train}}_t - y^{\text{train}}_{t-m}}},
    \label{eq:mase-def}
\end{equation}
where $y^{\text{train}}_{1:N}$ is the in-sample (training) portion of the same series and $m$ is the seasonal period (e.g., $m = 24$ for hourly data with daily seasonality, $m = 1$ for non-seasonal series). The denominator is the in-sample mean absolute error of the \emph{seasonal naive} predictor $\hat y_t^{\text{snaive}} = y_{t-m}$, providing a series-specific scaling that makes the metric unitless and comparable across series of different magnitudes. A MASE below $1$ means the model beats the seasonal-naive predictor on that series in mean absolute error; above $1$ means it does not.

\paragraph{Aggregation across a configuration.}
A GIFT-Eval ``configuration'' $c$ is a (dataset, prediction-horizon) pair. The benchmark evaluates the model on a fixed test split of multiple windows from each dataset; let $\mathcal{W}_c$ denote this set of test windows. The configuration-level raw MASE is the mean across windows:
\begin{equation}
    \mathrm{MASE}(\text{model}, c) \;=\; \frac{1}{\abs{\mathcal{W}_c}} \sum_{(\hat y, y) \in \mathcal{W}_c} \mathrm{MASE}(\hat y, y).
\end{equation}

\paragraph{Normalised MASE.}
Because absolute MASE values vary by orders of magnitude across configurations (some series are easy to forecast, some are not), the GIFT-Eval leaderboard divides each configuration's MASE by that configuration's seasonal-naive MASE before aggregating across configurations:
\begin{equation}
    \mathrm{MASE}^*(\text{model}, c) \;=\; \frac{\mathrm{MASE}(\text{model}, c)}{\mathrm{MASE}(\text{seasonal naive}, c)}.
    \label{eq:normalised-mase}
\end{equation}
A configuration-level $\mathrm{MASE}^*$ below $1$ means the model beats seasonal-naive on that configuration; above $1$ means it does not. Note that $\mathrm{MASE}^*$ is the ratio of the model's MASE to the seasonal-naive MASE, both computed using \eqref{eq:mase-def}; the seasonal-naive denominator therefore cancels, and $\mathrm{MASE}^*(\text{model}, c)$ is equivalently the ratio of the model's window-mean absolute error to the seasonal-naive predictor's window-mean absolute error on configuration $c$.

\paragraph{Geometric-mean aggregation.}
The headline aggregate is the geometric mean of $\mathrm{MASE}^*$ across all $C = 97$ configurations:
\begin{equation}
    \overline{\mathrm{MASE}^*}(\text{model}) \;=\; \left(\prod_{c=1}^{C} \mathrm{MASE}^*(\text{model}, c)\right)^{1/C} \;=\; \exp\!\left(\frac{1}{C}\sum_{c=1}^{C} \log \mathrm{MASE}^*(\text{model}, c)\right).
    \label{eq:geomean}
\end{equation}
The geometric mean is preferred over the arithmetic mean because it is symmetric in the log-domain ($\log\overline{\mathrm{MASE}^*}$ is the arithmetic mean of $\log\mathrm{MASE}^*$), so the aggregate cannot be dominated by a single configuration on which one model has an extreme MASE. Equivalently, it averages \emph{relative} performance across configurations. As a reference point, our retrained Moirai-2 Small \Baseline{} attains $\overline{\mathrm{MASE}^*} \approx 0.862$ at $10$K steps; the official Moirai-2.0-R-Small release attains $\approx 0.728$ on the same metric (Appendix~\ref{app:official-moirai}).

\paragraph{Reporting convention.}
We report cross-seed mean and standard deviation of $\overline{\mathrm{MASE}^*}$ together with per-configuration paired-comparison win counts (out of $97$ per seed; $485$ pooled across the $5$ seeds we use). Headline relative changes between methods are reported as $\Delta\% = \big(\overline{\mathrm{MASE}^*}(\text{treatment}) - \overline{\mathrm{MASE}^*}(\text{control})\big) / \overline{\mathrm{MASE}^*}(\text{control}) \cdot 100$, so a negative $\Delta\%$ means the treatment achieves lower MASE (an improvement). We do not report $p$-values for the per-configuration win counts: although a paired sign test would assign extremely small values to our pooled wins, the $5 \times 97 = 485$ paired observations are not independent (each of the $97$ configurations is reused across all $5$ seeds with the same training data and evaluation setup), so the iid-binomial null is mis-specified and any nominal $p$-value should be read with caution.

\subsection{The Polynomial Preconditioning Channel}
\label{sec:method}

We now describe the architectural modification that delivers the preconditioning channel of Eq.~\eqref{eq:residual-channel} into the model. Let $x \in \R^T$ be a univariate input series, normalised to zero mean and unit variance over the observed prefix. We form the residual
\[
    r_t \;:=\; \tilde T_d^*(B_s) x_t - x_t
\]
over the full sequence \emph{before} patching, then partition $r$ into patches $\mathbf{r}_i$ of length $P$ aligned with the target patches $\mathbf{p}_i$. The per-patch input projection is widened to accept the augmented triple $(\mathbf{p}_i, \mathbf{m}_i, \mathbf{r}_i)$:
\begin{equation}
    \mathbf{e}_i = \mathrm{ResBlock}_{3P \to d_{\text{model}}}\bigl([\mathbf{p}_i; \mathbf{m}_i; \mathbf{r}_i]\bigr) \in \R^{d_{\text{model}}},
    \label{eq:method-projection}
\end{equation}
where $\mathrm{ResBlock}$ consists of a widened linear layer followed by a SiLU nonlinearity, a second linear layer, and a residual connection. Only the first linear layer widens; the transformer stack, attention layers, output heads, and normalisation layers are unchanged.

\paragraph{Channel dropout.}
During training, each patch's preconditioning channel is independently zeroed with probability $p_{\text{drop}} = 0.1$; at inference the full preconditioning signal is always supplied. Channel dropout serves two purposes: (i) it prevents catastrophic dependence on the channel, since models trained without dropout degrade by $\sim 36\%$ in MASE when the channel is replaced by zeros at inference (Appendix~\ref{app:inference-ablation}); and (ii) it has a small regularising effect, empirically yielding a $\sim 0.3\%$ improvement over the no-dropout variant.

\paragraph{Parameter count.}
For Moirai-2 Small ($P = 16$, $d_{\text{model}} = 384$), widening the first linear layer from $32 \to 384$ (target + mask) to $48 \to 384$ (target + mask + preconditioning) adds $16 \cdot 384 = 6{,}144$ parameters; the residual path contributes another $\approx 6{,}000$. The total overhead is $\approx 12$K parameters, or $0.105\%$ of the $11.4$M baseline. No other parameters are added.

%==============================================================================
\section{Empirical Results}
\label{sec:empirical}
%==============================================================================

\subsection{Experimental Setup}

Our experimental protocol exactly matches Section~\ref{sec:patch-transformers}: we train Moirai-2 Small from scratch on LOTSA v1 (with the official Moirai-2 weighting) for $10$K optimizer steps---and $100$K in Section~\ref{sec:100k}---and evaluate on GIFT-Eval at context length $4{,}000$ across all $97$ dataset $\times$ horizon configurations. The four conditions of interest below share \emph{identical} training data, optimizer, schedule, seeds, and evaluation pipeline; only the content of the auxiliary input channel differs. We run five seeds per condition, $\{0, 1, 2, 7, 42\}$, yielding $5 \times 97 = 485$ paired comparisons per contrast. All headline aggregates use the geometric-mean normalised MASE of Eq.~\eqref{eq:normalised-mase}, with cross-seed mean and standard deviation reported throughout.

\subsection{Main Results}
\label{sec:main-results}

We compare four conditions, all with identical Moirai-2 Small architecture, training data, optimizer, and seeds; only the auxiliary input channel content differs. \emph{\HD{}} (ours) supplies the Chebyshev residual channel of Eq.~\eqref{eq:residual-channel}. The two capacity-matched controls test whether the gain is driven by the polynomial content or merely by the extra parameters in the widened input projection: \emph{\Zero{}} fills the auxiliary channel with zeros (the projection layer is widened identically to \HD{} but receives a constant input), while \emph{\Dup{}} fills it with a copy of the target patch (the projection sees the same information twice). Both controls have the same parameter count as \HD{} ($+12$K parameters over \Baseline{}). \emph{\Baseline{}} is unmodified Moirai-2 Small.

\begin{table}[h]
\centering
\caption{Main results on GIFT-Eval, 5 seeds, 485 paired comparisons per method. $\Delta\%$ is the geometric-mean MASE change relative to \Baseline{}; negative values mean lower MASE (improvement). Win rates are paired per-configuration comparisons. \Zero{} and \Dup{} are capacity-matched controls (identical architecture and parameter count to \HD{}; only the channel content differs).}
\label{tab:main}
\begin{tabular}{lccc}
\toprule
\textbf{Method} & \textbf{GeoMean MASE} & \textbf{$\Delta$\% vs \Baseline{}} & \textbf{Wins / 485} \\
\midrule
\HD{} (Ours) & $\mathbf{0.837}$ & $\mathbf{-2.89\%}$ & $\mathbf{329}$ \\
\Zero{} (capacity control) & $0.858$ & $-0.45\%$ & $262$ \\
\Baseline{} & $0.862$ & --- & --- \\
\Dup{} (capacity control) & $0.865$ & $+0.38\%$ & $217$ \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:main} shows that \HD{} improves geometric-mean MASE by $2.89\%$ over \Baseline{} at $10$K steps. Notably, neither capacity control provides a comparable benefit: \Dup{} \emph{hurts} performance ($44.7\%$ wins vs.\ $50\%$ expected by chance), and \Zero{} is at most weakly beneficial. These controls are diagnostic: they isolate the \emph{polynomial content} of the preconditioning channel, as opposed to the mere presence of an additional input channel or an extra $12$K parameters, as the source of the gain.

\subsection{Ablation: Filter Degree, Stride, Basis, and Dropout}

\begin{table}[h]
\centering
\small
\caption{Filter ablations on GIFT-Eval (97 configs, $10$K steps). Degree row uses 5 seeds (mean); other rows use seed 0. Best in \textbf{bold}.}
\label{tab:ablations-main}
\begin{tabular}{lccccccc}
\toprule
\multicolumn{8}{c}{\textbf{(a) Degree} (Chebyshev, $s=16$, dropout $10\%$, $5$ seeds, mean MASE)} \\
\midrule
$d$ & $2$ & $3$ & $\mathbf{4}$ & $5$ & $6$ & $7$ & $8$ \\
MASE & $0.850$ & $0.842$ & $\mathbf{0.837}$ & $0.842$ & $0.840$ & $0.847$ & $0.841$ \\
Std & $0.013$ & $0.005$ & $0.009$ & $0.008$ & $0.011$ & $0.011$ & $0.005$ \\
\midrule
\multicolumn{8}{c}{\textbf{(b) Stride} (Chebyshev $d{=}4$, dropout $10\%$, seed 0)} \\
\midrule
$s$ & $4$ & $8$ & $\mathbf{16}$ & $32$ & & & \\
MASE & $0.841$ & $0.852$ & $\mathbf{0.823}$ & $0.848$ & & & \\
\midrule
\multicolumn{8}{c}{\textbf{(c) Basis} ($d{=}4$, $s{=}16$, dropout $10\%$, seed 0)} \\
\midrule
Basis & \textbf{Cheby} & EMA & Legendre & FiniteDiff & & & \\
MASE & $\mathbf{0.823}$ & $0.833$ & $0.840$ & $0.857$ & & & \\
\midrule
\multicolumn{8}{c}{\textbf{(d) Channel Dropout} (Chebyshev $d{=}4$, $s{=}16$, seed 0)} \\
\midrule
$p_{\text{drop}}$ & $0\%$ & $2\%$ & $\mathbf{10\%}$ & $20\%$ & & & \\
MASE & $0.838$ & $0.839$ & $\mathbf{0.823}$ & $0.842$ & & & \\
\bottomrule
\end{tabular}
\end{table}

The ablations in Table~\ref{tab:ablations-main} align with theoretical predictions. (a) Across $d \in \{2, \ldots, 8\}$ at 5 seeds each, $d = 4$ has the lowest mean MASE, but the curve is shallow: $d \in \{3, 5, 6, 8\}$ are within roughly one seed-to-seed standard deviation, and only $d = 2$ (a single non-trivial lag) and $d = 7$ trail meaningfully. We treat $d=4$ as a representative working choice rather than a uniquely optimal one---the qualitative ``every $d \geq 2$ improves over the baseline'' conclusion is the seed-robust statement, and is what theory predicts. (b) Stride $s = P$ is optimal: filter taps align exactly with patch boundaries, so the preconditioning channel injects cross-patch structure at the natural scale. (c) Chebyshev beats EMA (a one-sided low-pass filter), Legendre (orthogonal but not minimax), and finite-difference (degenerate high-pass), confirming the minimax-optimality argument. (d) Dropout $10\%$ avoids both over-reliance on the channel (low dropout) and under-use (high dropout).

\subsection{FEV-Bench Cross-Benchmark Validation}
\label{sec:fevbench}

To verify that our findings generalize beyond GIFT-Eval, we evaluate on the independent \citet{das2024decoder} FEV-Bench (100 tasks). Following the same protocol as Section~\ref{sec:main-results}, we report normalized geomean MASE and FEV-Bench's scaled quantile loss (SQL):

\begin{table}[h]
\centering
\small
\caption{FEV-Bench results (5-seed mean for \HD{}, single-seed for \Baseline{}).}
\label{tab:fevbench}
\begin{tabular}{lccc}
\toprule
\textbf{Method} & \textbf{Geomean MASE} & \textbf{SQL} & \textbf{Wins/100} \\
\midrule
\HD{} (Ours) & $\mathbf{1.2532 \pm 0.0064}$ & $\mathbf{1.0236 \pm 0.0053}$ & $\mathbf{72}$ \\
\Baseline{} & $1.2908$ & $1.0562$ & --- \\
\midrule
$\Delta\%$ vs $\Baseline{}$ & $-2.91\%$ & $-3.09\%$ & --- \\
\bottomrule
\end{tabular}
\end{table}

The FEV-Bench improvements ($-2.91\%$ MASE, $-3.09\%$ SQL, $72/100$ wins) are comparable in magnitude to the GIFT-Eval improvement, and the per-degree sweep on FEV-Bench reproduces the GIFT-Eval pattern: every $d \geq 2$ beats baseline, the curve is shallow, and $d \in \{4, 6, 8\}$ are tied near the optimum. Two limitations are worth noting. First, the FEV-Bench baseline is a single-seed run while \HD{} is a 5-seed mean, so we cannot quantify cross-seed baseline variability on FEV-Bench. Second, the comparison shares training data (LOTSA) with the GIFT-Eval comparison; FEV-Bench is independent at the \emph{evaluation} level (different tasks, different metric) but not at the training-data level. We therefore present FEV-Bench as \emph{supportive cross-benchmark validation} rather than as fully independent evidence.

\subsection{Context-Length Robustness}

Across context lengths $\{512, 1000, 2000, 4000\}$, the relative improvement is approximately constant at $\sim 5\%$ (Appendix~\ref{app:context}). Notably, \HD{} at context $2000$ outperforms \Baseline{} at context $4000$ ($0.833$ vs.\ $0.867$), suggesting that input preconditioning can substitute for doubling the context window.

\subsection{Extended Training ($100$K Steps)}
\label{sec:100k}

\begin{table}[h]
\centering
\small
\caption{100K-step training, 5 seeds. The benefit grows with training budget; capacity controls actively hurt.}
\label{tab:100k}
\begin{tabular}{lcc}
\toprule
\textbf{Method} & \textbf{GeoMean MASE} & \textbf{$\Delta$\% vs \Baseline{}} \\
\midrule
\HD{} (Ours) & $\mathbf{0.844}$ & $\mathbf{-3.88\%}$ \\
\Baseline{} & $0.878$ & --- \\
\Dup{} & $0.886$ & $+0.86\%$ \\
\Zero{} & $0.894$ & $+1.73\%$ \\
\bottomrule
\end{tabular}
\end{table}

Two observations: the benefit of polynomial preconditioning \emph{grows} with training budget ($-3.9\%$ at $100$K vs.\ $-2.9\%$ at $10$K), and the capacity controls that were near-neutral at $10$K now actively hurt. The \Baseline{} also shows higher across-seed variance at $100$K (range $0.863$--$0.909$), consistent with late-training instability that the preconditioning channel appears to regularize.

%==============================================================================
\section{Related Work}
\label{sec:related}
%==============================================================================

\paragraph{Time series foundation models.}
Transformer-based universal forecasters have proliferated~\citep{woo2024unified, liu2024moirai2, das2024decoder, ansari2024chronos, goswami2024moment, auer2025tirex}. Recent work critically examines the zero-shot claims~\citep{zhang2025tsfmready}, arguing that simple PCA+linear baselines can be competitive; our method strengthens compact foundation models in exactly this small-model regime. Most architectural innovations in the space (rotary embeddings, attention variants, mixture-of-experts~\citep{liu2024moirai2}) target the transformer stack; input preconditioning is orthogonal, acting on the input representation before any attention layer.

\paragraph{Patching and overlap.}
Patch-based architectures~\citep{nie2023patchtst, dosovitskiy2021image} dominate TSFMs. PatchTST uses overlapping patches to partially mitigate the cross-patch bottleneck; our approach achieves a similar effect at lower cost (no FLOP increase, no parameter increase in the encoder) and is complementary (one could combine both).

\paragraph{Spectral methods in sequence modeling.}
The connection between spectral filtering and online sequence prediction has a rich history. \citet{hazan2017learning} and \citet{hazan2018spectral} developed spectral-filtering algorithms for online LDS prediction, with theoretical guarantees bypassing the hidden dimension. \citet{marsden2024universal} unify and generalize this line via polynomial preconditioning, which directly motivates our filter design.

\paragraph{Classical FIR filter theory.}
FIR filter design with Chebyshev polynomials is a classical topic~\citep{parks1972chebyshev, oppenheim1999discrete}; our contribution is to bring this to bear on the input representation of a modern transformer forecaster, rather than as a post-hoc smoother.

%==============================================================================
\section{Discussion and Limitations}
\label{sec:discussion}
%==============================================================================

\paragraph{When does preconditioning help?}
The spectral story predicts that datasets whose power spectrum aligns with the filter's attenuation minima benefit least, and those whose mass aligns with passbands benefit most. The cross-domain pattern (Appendix~\ref{app:domain}) is consistent with this prediction in a coarse sense: transport and energy domains ($-4$ to $-7\%$ MASE) have strong autocorrelation at the $32$- and $64$-sample lookback scales of the filter at $15$-minute resolution (corresponding to $8/16$-hour intra-day structure); weather and healthcare, dominated by much-longer or stochastic correlations, show essentially no improvement. We do not claim that this matches the precise filter-frequency calculation in Theorem~\ref{thm:freq-response}; the $32$- and $64$-step filter lags and the $\omega = 0, 1/32, 1/16, \ldots$ attenuation frequencies are distinct mathematical objects, and the cross-domain evidence cannot distinguish their respective contributions.

\paragraph{Model scale.}
We find that the benefit of input preconditioning is most pronounced at the small ($11.4$M) Moirai-2 scale; at the base ($46$M) scale, the improvement is seed-dependent and not consistently significant. This is consistent with a capacity-efficiency interpretation: when a model has enough parameters to discover the relevant cross-patch correlations via attention, the inductive bias of a fixed Chebyshev preconditioner adds less value.

\paragraph{Limitations.}
We focused on univariate, single-variate forecasting; extensions to multivariate and probabilistic forecasting are left as future work. The Chebyshev choice was validated by ablation against a handful of alternatives (Legendre, EMA, finite differences); exhaustive search over the monic-degree-$d$ family is computationally infeasible but could be approximated by closed-form optimal linear filters (Wiener filters) in a synthetic setting.

%==============================================================================
\section{Conclusion}
%==============================================================================

We studied polynomial input preconditioning for patch-based universal time series forecasters, a minimal architectural intervention that injects cross-patch temporal information via a fixed Chebyshev FIR filter at $0.11\%$ parameter overhead. Bridging the classical Chebyshev minimax theorem, the recent framework of universal sequence preconditioning~\citep{marsden2024universal}, and modern patch-based transformer forecasters, we established a closed-form frequency response for our specific filter, proved a BIBO-stability lemma, and evaluated the method on the GIFT-Eval benchmark ($-2.9\%$ MASE at $10$K steps, growing to $-3.9\%$ at $100$K steps; $329$ of $485$ paired per-configuration wins). Capacity-matched controls confirmed that the improvement is driven by the polynomial content of the preconditioning channel rather than by extra parameters, and a per-domain breakdown is consistent with the prediction that cross-dataset variation in improvement aligns with dataset-level frequency structure.

\newpage
\bibliographystyle{plainnat}
\bibliography{references}

\newpage
\appendix

\section{Detailed Methodology}
\label{app:methodology}

\subsection{Training Protocol}

All experiments use identical training infrastructure on NVIDIA H100/A100 GPUs.

\begin{center}
\small
\begin{tabular}{ll}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Architecture & Moirai-2 Small (6 layers, $d=384$, $d_{\text{ff}}=1024$, $P=16$) \\
Total parameters & 11.4M (Baseline), 11.4M $+$ 12K (\HD{}) \\
Training data & LOTSA v1, Moirai-2 config (weighted, variate-proportional) \\
Steps (10K runs) & 100 epochs $\times$ 100 batches $=$ 10{,}000 steps \\
Steps (100K runs) & 1{,}000 epochs $\times$ 100 batches $=$ 100{,}000 steps \\
Batch size & 256 \\
Optimizer & AdamW ($\beta_1=0.9$, $\beta_2=0.98$, $\epsilon=10^{-6}$, wd $=0.1$) \\
LR schedule & Cosine annealing, peak $\eta = 10^{-3}$, min $\eta = 0$ \\
Warmup & 1{,}000 steps (linear) \\
Precision & bf16 mixed precision \\
Anomaly filtering & $z$-score threshold $=8.0$, variance ratio $=0.0$ (disabled) \\
Random seeds & 0, 1, 2, 7, 42 (5 seeds for main results) \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Evaluation Protocol}

All evaluations use the GIFT-Eval benchmark \citep{vijay2024gifteval} with 97 dataset$\times$horizon configurations:
\begin{itemize}[leftmargin=1.5em]
    \item Context length: 4{,}000 tokens (maximum supported)
    \item Patch size: 32 for most frequencies, 8 for quarterly (following Moirai conventions)
    \item Batch size: 64
    \item Metric: Normalized MASE (each configuration's MASE divided by the seasonal naive baseline's MASE)
    \item Aggregation: geometric mean across 97 configurations
    \item Reported quantities: cross-seed mean and standard deviation, per-seed MASE, and per-configuration win counts (out of 97 per seed; $5 \times 97 = 485$ when pooled across seeds)
\end{itemize}

\subsection{Preconditioning Channel Implementation}

The Chebyshev polynomial $T_d(z)$ is expanded in the monomial basis to obtain FIR filter coefficients $c_0, \ldots, c_d$. The preconditioned signal for time step $t$ is
\[
    r_t \;=\; \left(\sum_{k=0}^{d} c_k \, x_{t-ks}\right) - x_t,
\]
computed over the full (normalized, zero-mean unit-variance) time series before patching. Unobserved time steps are masked to zero. During training with channel dropout $p$, each patch's preconditioning channel independently has probability $p$ of being replaced with zeros. At inference, the full preconditioning channel is always provided.

The preconditioning channel is concatenated with the target and mask channels before the input projection. The input projection is the only architectural change relative to Baseline; all downstream layers (transformer encoder, output projection) are identical.

%==============================================================================
\section{Capacity Controls}
\label{app:capacity}
%==============================================================================

The Zero and Duplicate conditions serve as capacity controls: they have the \emph{same architecture and parameter count} as \HD{}, differing only in the content of the extra channel.

\begin{table}[h]
\centering
\caption{Capacity control experiments at 10K steps (5 seeds, $\eta = 10^{-3}$). Same architecture width, different channel content.}
\label{tab:capacity-10k}
\small
\begin{tabular}{lcccc}
\toprule
\textbf{Method} & \textbf{Content} & \textbf{Mean MASE} & \textbf{Std} & \textbf{Wins/485} \\
\midrule
\textbf{\HD} & Chebyshev $\tilde T_4(B^{16})$ & \textbf{0.8368} & 0.0083 & \textbf{329} \\
Zero & All zeros & 0.8578 & 0.0102 & 262 \\
Baseline & No extra channel & 0.8617 & 0.0062 & --- \\
Duplicate & Copy of input & 0.8650 & 0.0081 & 217 \\
\bottomrule
\end{tabular}
\end{table}

\begin{itemize}[leftmargin=1.5em]
    \item \textbf{Duplicate channel hurts}: win rate $44.7\%$, \emph{worse} than the $50\%$ expected by chance.
    \item \textbf{Zero channel is neutral}: win rate $54.0\%$, tiny effect ($-0.45\%$).
    \item \HD{} vs.\ Duplicate: $336/485$ wins (\HD{} beats Duplicate on every seed).
    \item \HD{} vs.\ Zero: $323/485$ wins (\HD{} beats Zero on every seed).
\end{itemize}

At 100K steps, both controls actively hurt: Duplicate degrades by $0.9\%$ and Zero by $1.7\%$ relative to Baseline. Full table in Appendix~\ref{app:100k}.

%==============================================================================
\section{Filter Ablations}
\label{app:ablations}
%==============================================================================

All ablations use seed 0 with $10\%$ channel dropout unless otherwise noted; the polynomial-degree sweep is the exception and is reported over $5$ seeds each.

\subsection{Polynomial Degree (Chebyshev, $s=16$, $10\%$ dropout)}

\begin{table}[h]
\centering
\small
\begin{tabular}{lccc}
\toprule
\textbf{Degree} & \textbf{Residual lags} & \textbf{MASE (mean $\pm$ std, $n=5$)} & \textbf{$\Delta\%$ vs.\ BL} \\
\midrule
2 & 1 (at $2P$) & $0.8502 \pm 0.0125$ & $-1.33\%$ \\
3 & 1 (at $2P$) & $0.8422 \pm 0.0053$ & $-2.27\%$ \\
\textbf{4} & 2 (at $2P$, $4P$) & $\mathbf{0.8368 \pm 0.0092}$ & $\mathbf{-2.89\%}$ \\
5 & 2 (at $2P$, $4P$) & $0.8423 \pm 0.0078$ & $-2.25\%$ \\
6 & 3 (at $2P$, $4P$, $6P$) & $0.8400 \pm 0.0105$ & $-2.52\%$ \\
7 & 3 (at $2P$, $4P$, $6P$) & $0.8470 \pm 0.0110$ & $-1.71\%$ \\
8 & 4 (at $2P$, $4P$, $6P$, $8P$) & $0.8414 \pm 0.0050$ & $-2.36\%$ \\
\midrule
Baseline & --- & $0.8617 \pm 0.0062$ & --- \\
\bottomrule
\end{tabular}
\end{table}

Every $d \geq 2$ beats Baseline. The mean curve is shallow: $d=4$ achieves the lowest geomean MASE but $d \in \{3, 5, 6, 8\}$ are within roughly one seed-to-seed standard deviation of it. The extreme $d=2$ has only one non-trivial residual lag (at $2P$). Note that under the reciprocal-monic convention, the residual filter for any odd $d$ contains only even-lag terms, so the lag structure of $d \in \{3, 5, 7\}$ matches that of $d-1 \in \{2, 4, 6\}$ respectively, with different coefficients; this can be verified by direct expansion of $\tilde T_d^*(z) - 1$.

\subsection{Filter Stride (Chebyshev $d=4$, $10\%$ dropout)}

\begin{table}[h]
\centering
\small
\begin{tabular}{ccccc}
\toprule
\textbf{Stride} & \textbf{Lookback} & \textbf{MASE} & \textbf{$\Delta\%$ vs.\ BL} & \textbf{Wins/97} \\
\midrule
\textbf{16 ($=P$)} & 32, 64 steps & \textbf{0.8234} & $\mathbf{-4.99\%}$ & \textbf{66} \\
4 & 8, 16 steps & 0.8406 & $-3.00\%$ & 57 \\
8 & 16, 32 steps & 0.8518 & $-1.71\%$ & 62 \\
32 & 64, 128 steps & 0.8484 & $-2.10\%$ & 63 \\
8+16 (dual) & 16, 32, 64 steps & 0.8389 & $-3.20\%$ & 62 \\
\bottomrule
\end{tabular}
\end{table}

Stride $s = 16 = P$ is optimal: filter taps align exactly with patch boundaries, providing maximally informative cross-patch lookback.

\subsection{Filter Basis ($d=4$, $s=16$, $10\%$ dropout)}

\begin{table}[h]
\centering
\small
\begin{tabular}{lcccc}
\toprule
\textbf{Basis} & \textbf{MASE} & \textbf{$\Delta\%$} & \textbf{Wins vs.\ BL} & \textbf{Wins vs.\ Cheby} \\
\midrule
\textbf{Chebyshev} $\tilde T_4^*$ & \textbf{0.8234} & $\mathbf{-4.99\%}$ & 66/97 & --- \\
EMA & 0.8329 & $-3.89\%$ & 65/97 & 41/97 \\
Legendre $\tilde P_4^*$ & 0.8400 & $-3.07\%$ & 62/97 & 30/97 \\
Legendre $\tilde P_6^*$ & 0.8397 & $-3.10\%$ & 64/97 & 50/97 \\
Finite differences & 0.8569 & $-1.12\%$ & 54/97 & 29/97 \\
Duplicate & 0.8583 & $-0.96\%$ & 42/97 & 27/97 \\
\bottomrule
\end{tabular}
\label{tab:basis-ablation}
\end{table}

The basis ordering Chebyshev $>$ EMA $>$ Legendre $>$ finite differences confirms the theoretical prediction of Section~\ref{sec:basis-comparison}: the family that minimizes the $L^\infty$ norm on $[-1,1]$ outperforms families optimised for other norms.

\subsection{Channel Dropout Rate (Chebyshev $d=4$, $s=16$)}

\begin{table}[h]
\centering
\small
\begin{tabular}{cccc}
\toprule
\textbf{Dropout} & \textbf{MASE} & \textbf{$\Delta\%$ vs.\ BL} & \textbf{Wins/97} \\
\midrule
$0\%$ (\Dnodrop)$^\dagger$ & 0.8378 & $-3.50\%$ & --- \\
$2\%$ & 0.8388 & $-3.20\%$ & 63 \\
$3\%$ & 0.8497 & $-1.95\%$ & 55 \\
\textbf{$10\%$ (\HD{})} & \textbf{0.8234} & $\mathbf{-4.99\%}$ & \textbf{66} \\
$20\%$ & 0.8422 & $-2.81\%$ & 67 \\
$30\%$ & 0.8559 & $-1.23\%$ & 61 \\
\bottomrule
\end{tabular}\\[2pt]
{\footnotesize $^\dagger$\Dnodrop{} MASE is the 5-seed mean. \Dnodrop{} and \HD{} are essentially tied; dropout helps marginally.}
\end{table}

\subsection{Learned FIR Coefficients (seed 0)}

\begin{table}[h]
\centering
\small
\begin{tabular}{lccc}
\toprule
\textbf{Initialization} & \textbf{Learned coefficients} & \textbf{MASE} & \textbf{$\Delta\%$ vs.\ BL} \\
\midrule
Chebyshev init & $[-0.026, -1.032, -0.123, 0.007]$ & 0.8332 & $-3.85\%$ \\
Zero init & $[0.308, 0.177, 0.057, 0.045]$ & 0.8358 & $-3.57\%$ \\
\textbf{Fixed Chebyshev} & $[0, -1, 0, 0.125]$ (not learned) & \textbf{0.8234} & $\mathbf{-4.99\%}$ \\
\bottomrule
\end{tabular}
\label{tab:learned-fir}
\end{table}

The two learned filters converge to distinct basins (high-pass vs.\ smoothing). Both beat Baseline but are slightly worse than fixed Chebyshev (51/97 head-to-head wins for fixed). Fixing the polynomial provides an inductive bias the model would otherwise have to discover.

\subsection{Latent Preconditioning (seed 0)}

Applying the polynomial filter \emph{after} the input projection (in latent space) rather than before patching:

\begin{table}[h]
\centering
\small
\begin{tabular}{lccc}
\toprule
\textbf{Method} & \textbf{Filter location} & \textbf{MASE} & \textbf{vs.\ BL} \\
\midrule
\HD{} (input-level) & Before patching & 0.8234 & $-4.99\%$ \\
Latent $s=1$ & After input projection & 1.1538 & $+33.1\%$ \\
Latent $s=4$ & After input projection & 1.2243 & $+41.3\%$ \\
\bottomrule
\end{tabular}
\end{table}

Latent preconditioning is catastrophic. The filter must operate on the raw time series, not on learned embeddings: the spectral structure is destroyed by the input projection's nonlinear mixing.

%==============================================================================
\section{Polynomial Coefficients}
\label{app:poly-coeffs}
%==============================================================================

This appendix lists the explicit FIR coefficients for every polynomial basis discussed in this paper. Throughout, we use the convention of Section~\ref{sec:precondition}: the residual filter applied to the input is $r_t = (p(B_s) x)_t - x_t = \sum_{k=1}^{d} c_k x_{t-ks}$, where $c_1, \ldots, c_d$ are the non-leading coefficients of the monic polynomial $p$ ordered so that $c_k$ is the coefficient of $z^{d-k}$ in $z^d p(1/z)$ (the reciprocal monic polynomial). The default stride is $s = P = 16$.

\paragraph{Monic Chebyshev (first kind), $\tilde T_d(x) = T_d(x)/2^{d-1}$.}
The classical Chebyshev polynomials satisfy the recurrence $T_0 = 1$, $T_1(x) = x$, $T_{d+1}(x) = 2x T_d(x) - T_{d-1}(x)$. After dividing by the leading coefficient $2^{d-1}$, the monic forms are obtained explicitly. The reciprocal-monic FIR coefficients $(c_1, \ldots, c_d)$ at stride $s$ are listed in Table~\ref{tab:cheb-coeffs}.

\begin{table}[h]
\centering
\caption{Monic Chebyshev (first kind): polynomial coefficients $\tilde T_d(x)$ in the monomial basis (descending powers), and the corresponding residual FIR coefficients $(c_1, \ldots, c_d)$ at stride $s = P = 16$. Effective lookback in time-step units is $ks$ for the $k$-th nonzero tap. Boldface marks the default $d = 4$ filter used by \HD{} and \texttt{d4}.}
\label{tab:cheb-coeffs}
\small
\begin{tabular}{rlcl}
\toprule
$d$ & $\tilde T_d(x)$ & FIR $(c_1, \ldots, c_d)$ & Residual $r_t$ at $s = 16$ \\
\midrule
$2$ & $x^2 - \tfrac{1}{2}$                                       & $(0, -\tfrac{1}{2})$                            & $-\tfrac{1}{2} x_{t-32}$ \\
$3$ & $x^3 - \tfrac{3}{4} x$                                     & $(0, -\tfrac{3}{4}, 0)$                         & $-\tfrac{3}{4} x_{t-32}$ \\
$\mathbf{4}$ & $x^4 - x^2 + \tfrac{1}{8}$                        & $\mathbf{(0, -1, 0, \tfrac{1}{8})}$              & $\mathbf{-x_{t-32} + \tfrac{1}{8} \, x_{t-64}}$ \\
$5$ & $x^5 - \tfrac{5}{4} x^3 + \tfrac{5}{16} x$                 & $(0, -\tfrac{5}{4}, 0, \tfrac{5}{16}, 0)$       & $-\tfrac{5}{4} x_{t-32} + \tfrac{5}{16} x_{t-64}$ \\
$6$ & $x^6 - \tfrac{3}{2} x^4 + \tfrac{9}{16} x^2 - \tfrac{1}{32}$ & $(0, -\tfrac{3}{2}, 0, \tfrac{9}{16}, 0, -\tfrac{1}{32})$ & $-\tfrac{3}{2} x_{t-32} + \tfrac{9}{16} x_{t-64} - \tfrac{1}{32} x_{t-96}$ \\
$7$ & $x^7 - \tfrac{7}{4} x^5 + \tfrac{14}{16} x^3 - \tfrac{7}{64} x$ & $(0, -\tfrac{7}{4}, 0, \tfrac{14}{16}, 0, -\tfrac{7}{64}, 0)$ & $-\tfrac{7}{4} x_{t-32} + \tfrac{7}{8} x_{t-64} - \tfrac{7}{64} x_{t-96}$ \\
$8$ & $x^8 - 2 x^6 + \tfrac{5}{4} x^4 - \tfrac{1}{4} x^2 + \tfrac{1}{128}$ & $(0, -2, 0, \tfrac{5}{4}, 0, -\tfrac{1}{4}, 0, \tfrac{1}{128})$ & $-2 x_{t-32} + \tfrac{5}{4} x_{t-64} - \tfrac{1}{4} x_{t-96} + \tfrac{1}{128} x_{t-128}$ \\
\bottomrule
\end{tabular}
\end{table}

For odd $d$, the residual filter at stride $s$ has a tap at $x_{t-0} = x_t$ that cancels with the identity term, so the residual contains only \emph{even} lags $2s, 4s, \ldots$, regardless of polynomial parity in $x$. For example, the $d=3$ residual at $s = 16$ is supported only at lag $32$; the $d=5$ residual at $s = 16$ is supported at lags $32$ and $64$.

\paragraph{Monic Chebyshev second kind, $\tilde U_d(x) = U_d(x)/2^d$.}
Defined by $U_0 = 1$, $U_1(x) = 2x$, $U_{d+1}(x) = 2x U_d(x) - U_{d-1}(x)$. The first few monic forms and corresponding residual FIRs at $s = 16$ are:
\[
\begin{aligned}
\tilde U_2(x) &= x^2 - \tfrac{1}{4}, &\quad r_t &= -\tfrac{1}{4} x_{t-32}, \\
\tilde U_3(x) &= x^3 - \tfrac{1}{2} x, &\quad r_t &= -\tfrac{1}{2} x_{t-32}, \\
\tilde U_4(x) &= x^4 - \tfrac{3}{4} x^2 + \tfrac{1}{16}, &\quad r_t &= -\tfrac{3}{4} x_{t-32} + \tfrac{1}{16} x_{t-64}.
\end{aligned}
\]
The $\ell_\infty$ norm of monic $\tilde U_d$ on $[-1,1]$ is $1/2^{d-1}$ \citep[Eq.~1.41]{mason2003chebyshev}, twice that of $\tilde T_d$, so the spectral attenuation predicted by the equioscillation argument is half as strong as for the first kind.

\paragraph{Monic Legendre, $\tilde P_d(x) = P_d(x) \cdot 2^d (d!)^2 / (2d)!$.}
Defined by Bonnet's recurrence $(d+1) P_{d+1}(x) = (2d+1) x P_d(x) - d P_{d-1}(x)$, $P_0 = 1$, $P_1(x) = x$. The leading coefficient of $P_d$ is $(2d)!/(2^d (d!)^2)$, so dividing by it yields the monic form. The first few monic Legendre polynomials and residual FIRs at $s = 16$ are:
\[
\begin{aligned}
\tilde P_2(x) &= x^2 - \tfrac{1}{3}, &\quad r_t &= -\tfrac{1}{3} x_{t-32}, \\
\tilde P_3(x) &= x^3 - \tfrac{3}{5} x, &\quad r_t &= -\tfrac{3}{5} x_{t-32}, \\
\tilde P_4(x) &= x^4 - \tfrac{6}{7} x^2 + \tfrac{3}{35}, &\quad r_t &= -\tfrac{6}{7} x_{t-32} + \tfrac{3}{35} x_{t-64}.
\end{aligned}
\]
The $\ell_\infty$ norm of monic $\tilde P_d$ on $[-1,1]$ is strictly larger than that of $\tilde T_d$ for all $d \geq 2$: their ratio is $2 \binom{2d}{d}/4^d \to 2/\sqrt{\pi d}$ as $d \to \infty$ (Proposition~\ref{prop:cheb-vs-legendre}). Legendre minimises the $\ell_2$ norm $\bigl(\int_{-1}^{1} p(x)^2 \, dx\bigr)^{1/2}$ rather than the $\ell_\infty$ norm.

\paragraph{Other bases used in ablations.}
\begin{itemize}[leftmargin=1.5em]
    \item \textbf{Finite-difference} (degree $d$): $p(z) = (z - 1)^d$. For $d = 4$ at $s = 16$ this expands to $r_t = -4 x_{t-16} + 6 x_{t-32} - 4 x_{t-48} + x_{t-64}$. Heavy high-pass behaviour, no spectral-attenuation guarantee.
    \item \textbf{Exponential moving average} (EMA, half-life $\tau$): a one-sided low-pass filter with infinite impulse response coefficients $a_k = (1-\alpha) \alpha^k$, truncated to the window we use. Not polynomial-based.
    \item \textbf{Learned FIR} (no fixed basis): the model learns the $d = 4$ residual coefficients $(c_1, c_2, c_3, c_4) \in \R^4$ from a Gaussian initialisation jointly with the rest of the network. The learned solution depends on initialisation (Appendix~\ref{app:ablations}, Table~\ref{tab:learned-fir}).
\end{itemize}

\paragraph{Multi-scale variants.}
The MS46 and MSHD10 variants stack two polynomial residual channels:
\begin{itemize}[leftmargin=1.5em]
    \item \texttt{MS46}: Chebyshev $d = 4$ at $s = 16$ \emph{and} Chebyshev $d = 6$ at $s = 16$, both stripped of the identity term. The two residuals $r^{(4)}, r^{(6)}$ are concatenated as separate channels alongside the target and mask, so the per-patch input projection widens from $2P \to d_{\text{model}}$ to $4P \to d_{\text{model}}$.
    \item \texttt{MSHD10}: same as \texttt{MS46} with $10\%$ per-patch channel dropout applied independently to each residual channel during training.
\end{itemize}

%==============================================================================
\section{Domain and Frequency Analysis}
\label{app:domain}
%==============================================================================

\begin{table}[h]
\centering
\caption{Domain analysis (\HD{} vs.\ BL, 5 seeds pooled, 485 observations).}
\small
\begin{tabular}{lccc}
\toprule
\textbf{Domain} & \textbf{$n$} & \textbf{$\Delta\%$} & \textbf{Wins} \\
\midrule
Demographics & 15 & $-9.83\%$ & 13/15 \\
Transport & 55 & $-6.94\%$ & 47/55 \\
Energy & 80 & $-3.97\%$ & 56/80 \\
M4 Competition & 30 & $-3.83\%$ & 23/30 \\
IT/Cloud & 120 & $-2.23\%$ & 81/120 \\
ETT (Industrial) & 80 & $-1.50\%$ & 54/80 \\
Retail/Sales & 20 & $-0.85\%$ & 14/20 \\
Weather & 75 & $-0.34\%$ & 36/75 \\
Healthcare & 10 & $-0.24\%$ & 5/10 \\
\bottomrule
\end{tabular}
\end{table}

The benefit is consistent with the mechanism, although we caution that the relevant filter-level mechanism has two distinct manifestations: (a) the residual filter's \emph{lookback lags}, which are $2s = 32$ and $4s = 64$ time steps (this controls which past samples enter the residual channel directly); and (b) the residual filter's \emph{attenuation/passband frequencies}, which are at $\omega = 0, 1/32, 1/16, \ldots$ cycles per sample (this controls which frequencies the residual channel preserves vs.\ attenuates). The two are mathematically distinct objects and need not coincide with dataset-level periodicities. For $15$-minute data, the $32/64$-step lookbacks correspond to $8/16$ hours of context---useful for transport and energy demand, which exhibit autocorrelation across that range---while the filter's attenuation periods of $\infty, 8\,\text{h}, 4\,\text{h}, \ldots$ are a distinct set. For daily weather data, both manifestations point to $32$/$64$-day scales which are too coarse to capture weekly/seasonal patterns. The cross-domain pattern is therefore consistent with both lookback and frequency interpretations, but the data does not let us distinguish their contributions.

\begin{table}[h]
\centering
\caption{Frequency analysis (\HD{} vs.\ BL, 5 seeds pooled).}
\small
\begin{tabular}{lccc}
\toprule
\textbf{Frequency} & \textbf{$n$} & \textbf{$\Delta\%$} & \textbf{Wins} \\
\midrule
Sub-hourly & 180 & $-3.40\%$ & 121/180 \\
Hourly & 155 & $-3.80\%$ & 113/155 \\
Daily & 75 & $-0.76\%$ & 46/75 \\
Weekly & 40 & $-0.73\%$ & 29/40 \\
Monthly & 25 & $-3.41\%$ & 13/25 \\
Quarterly & 5 & $-0.24\%$ & 3/5 \\
Yearly & 5 & $-5.92\%$ & 4/5 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Prediction horizon.} The benefit increases with prediction length: $-1.4\%$ for short horizons to $-5.1\%$ for long horizons (more negative = larger MASE reduction). Longer horizons require more temporal structure, which the preconditioning channel provides.

%==============================================================================
\section{Horizon Analysis: \Dnodrop{} vs.\ \HD{}}
\label{app:horizon}
%==============================================================================

The preconditioning filter for \Dnodrop{}/\HD{} ($d=4$, $s=16$) reduces to $r_t = -x_{t-32} + (1/8)\, x_{t-64}$. We test whether \HD{}'s channel-dropout regularisation buys robustness on long horizons that \Dnodrop{} lacks.

\begin{table}[h]
\centering
\caption{\Dnodrop{} vs.\ \HD{} vs.\ Baseline by GIFT-Eval term.}
\small
\begin{tabular}{lccccccc}
\toprule
\textbf{Bin} & \textbf{$n$} & \textbf{BL} & \textbf{\Dnodrop} & \textbf{\HD} & \textbf{\Dnodrop{} $\Delta\%$} & \textbf{\HD{} $\Delta\%$} & \textbf{\Dnodrop{} better} \\
\midrule
short  & 55 & 0.8403 & \textbf{0.8252} & 0.8282 & $-1.79\%$ & $-1.44\%$ & $24/55$ \\
medium & 21 & 0.8592 & 0.8301 & \textbf{0.8231} & $-3.39\%$ & $-4.20\%$ & $9/21$ \\
long   & 21 & 0.9237 & 0.8802 & \textbf{0.8744} & $-4.71\%$ & $-5.34\%$ & $7/21$ \\
\midrule
ALL    & 97 & 0.8618 & 0.8379 & \textbf{0.8368} & $-2.77\%$ & $-2.90\%$ & $40/97$ \\
\bottomrule
\end{tabular}
\end{table}

On short horizons \Dnodrop{} and \HD{} are within $0.4\%$ of each other; on medium and long horizons (PL $\geq 480$) \HD{} beats \Dnodrop{} by $0.7$--$0.85\%$ relative MASE---the direction the dropout-as-mismatch-regulariser story predicts but small relative to the $\sim 4\%$ HD-vs-Baseline gap.

%==============================================================================
\section{Context-Length Analysis}
\label{app:context}
%==============================================================================

\begin{table}[h]
\centering
\caption{Context-length analysis (seed 0). The benefit is approximately constant ($\sim 5\%$) across context lengths.}
\small
\begin{tabular}{ccccc}
\toprule
\textbf{Context} & \textbf{BL} & \textbf{\HD} & \textbf{$\Delta\%$} & \textbf{Wins/97} \\
\midrule
512    & 0.9432 & 0.8986 & $-4.73\%$ & 62 \\
1{,}000  & 0.9166 & 0.8706 & $-5.02\%$ & 65 \\
2{,}000  & 0.8813 & 0.8331 & $-5.47\%$ & 71 \\
4{,}000  & 0.8666 & 0.8234 & $-4.99\%$ & 66 \\
\bottomrule
\end{tabular}
\end{table}

\HD{} at context length 2{,}000 (MASE $0.8331$) outperforms Baseline at context length 4{,}000 (MASE $0.8666$), suggesting that input preconditioning can substitute for doubling the context window.

%==============================================================================
\section{Multi-Scale Variants}
\label{app:multiscale}
%==============================================================================

We tested two multi-scale variants providing two preconditioning channels (degree 4 + degree 6):

\begin{table}[h]
\centering
\small
\begin{tabular}{lcccc}
\toprule
\textbf{Method} & \textbf{Mean} & \textbf{Std} & \textbf{vs.\ BL} & \textbf{Wins/485} \\
\midrule
\HD{} ($d=4$, $10\%$ drop) & 0.8368 & 0.0083 & $-2.89\%$ & 329 \\
MS46 ($d=4{+}6$, $0\%$ drop) & 0.8434 & 0.0123 & $-2.13\%$ & 330 \\
MSHD10 ($d=4{+}6$, $10\%$ drop) & 0.8437 & 0.0072 & $-2.09\%$ & 315 \\
\bottomrule
\end{tabular}
\end{table}

All three methods are essentially tied head-to-head. MSHD10 has the lowest cross-seed standard deviation. We chose \HD{} as the headline configuration because it has the simplest architecture (single channel) while matching the multi-scale variants' performance.

%==============================================================================
\section{100K Extended Training}
\label{app:100k}
%==============================================================================

\begin{table}[h]
\centering
\caption{Full 100K training budget with capacity controls (5 seeds). \HD{} vs.\ BL: $-3.9\%$; \HD{} vs.\ Duplicate: $-4.7\%$; \HD{} vs.\ Zero: $-5.5\%$.}
\small
\begin{tabular}{lcccccc}
\toprule
& \textbf{s0} & \textbf{s1} & \textbf{s2} & \textbf{s7} & \textbf{s42} & \textbf{Mean} \\
\midrule
Baseline & 0.9092 & 0.8920 & 0.8633 & 0.8641 & 0.8636 & 0.8784 \\
\textbf{\HD} & \textbf{0.8349} & \textbf{0.8402} & 0.8586 & \textbf{0.8450} & \textbf{0.8430} & \textbf{0.8443} \\
MSHD10 & 0.8266 & --- & 0.8432 & 0.8437 & --- & 0.8378$^\dagger$ \\
Duplicate & 0.8887 & 0.8740 & 0.8974 & 0.8794 & 0.8903 & 0.8860 \\
Zero & 0.8917 & 0.8585 & 0.8750 & 0.8825 & 0.9610 & 0.8937 \\
\midrule
\HD{} $\Delta\%$ vs.\ BL & $-8.2\%$ & $-5.8\%$ & $-0.5\%$ & $-2.2\%$ & $-2.4\%$ & $-3.9\%$ \\
\bottomrule
\end{tabular}\\[2pt]
{\footnotesize $^\dagger$MSHD10 mean over 3 available seeds (0, 2, 7).}
\end{table}

Two findings stand out. First, the benefit of polynomial preconditioning \emph{grows} with training budget: \HD{} improves by $3.9\%$ at 100K (vs.\ $2.9\%$ at 10K). Second, the capacity controls that were near-neutral at 10K now actively hurt: Duplicate degrades by $0.9\%$ and Zero by $1.7\%$. Baseline shows higher variance across seeds at 100K (seed 0 reaches $0.9092$ due to late-training instability), while \HD{} remains stable (range $0.83$--$0.86$).

%==============================================================================
\section{Inference-Time Channel-Content Ablation}
\label{app:inference-ablation}
%==============================================================================

A model trained with Chebyshev preconditioning (\HD{}) is evaluated with different channel content at inference time:

\begin{table}[h]
\centering
\caption{Inference-time content ablation (seed 0, \HD{} model). The ordering confirms the model has learned to exploit specific polynomial structure, not generic extra capacity.}
\small
\begin{tabular}{lcc}
\toprule
\textbf{Inference channel content} & \textbf{MASE} & \textbf{vs.\ Normal} \\
\midrule
Normal (Chebyshev) & 0.8234 & --- \\
No preconditioning channel (Baseline arch) & 0.8666 & $+5.2\%$ \\
Zeros & 0.9003 & $+9.3\%$ \\
Random Gaussian noise & 0.9427 & $+14.5\%$ \\
Duplicate (copy of input) & 1.0699 & $+29.9\%$ \\
\bottomrule
\end{tabular}
\end{table}

The ordering Normal $\gg$ Baseline $>$ Zero $>$ Random $>$ Duplicate confirms that the model exploits the specific polynomial structure. Random and duplicate channels are worse than no preconditioning at all, demonstrating that the model does not simply use the extra channel for generic capacity.

%==============================================================================
\section{FEV-Bench Cross-Benchmark Validation}
\label{app:fevbench}
%==============================================================================

To verify that our findings generalize beyond GIFT-Eval, we evaluate on FEV-Bench \citep{das2024decoder} ($100$ tasks), an independent time-series forecasting benchmark, and reproduce the polynomial-degree sweep.

\begin{table}[h]
\centering
\caption{FEV-Bench polynomial-degree sweep, 5 seeds per degree. $\Delta\%$ relative to the single-seed baseline ($1.2908$).}
\small
\begin{tabular}{cccc}
\toprule
\textbf{Degree} & \textbf{MASE (mean $\pm$ std)} & \textbf{$\Delta\%$ vs.\ BL} & \textbf{SQL (mean $\pm$ std)} \\
\midrule
2 & $1.2780 \pm 0.0039$ & $-0.99\%$ & $1.0425 \pm 0.0032$ \\
3 & $1.2609 \pm 0.0173$ & $-2.32\%$ & $1.0288 \pm 0.0125$ \\
\textbf{4} & $\mathbf{1.2532 \pm 0.0064}$ & $\mathbf{-2.92\%}$ & $\mathbf{1.0236 \pm 0.0053}$ \\
5 & $1.2639 \pm 0.0069$ & $-2.08\%$ & $1.0309 \pm 0.0063$ \\
6 & $1.2521 \pm 0.0068$ & $-3.00\%$ & $1.0236 \pm 0.0060$ \\
7 & $1.2564 \pm 0.0114$ & $-2.66\%$ & $1.0263 \pm 0.0094$ \\
8 & $1.2518 \pm 0.0082$ & $-3.02\%$ & $1.0228 \pm 0.0059$ \\
\bottomrule
\end{tabular}
\end{table}

\HD{} ($d=4$) improves geometric-mean MASE by $2.91\%$ on FEV-Bench (5-seed mean $1.2532 \pm 0.0064$ vs.\ Baseline $1.2908$) and SQL (Salesforce's scaled quantile loss) by $3.09\%$ ($1.0236 \pm 0.0053$ vs.\ $1.0562$). The FEV-Bench degree sweep reproduces the qualitative GIFT-Eval pattern: every $d \geq 2$ beats Baseline, the curve is shallow, and $d \in \{4, 6, 8\}$ are statistically tied at the optimum.

%==============================================================================
\section{Comparison to Official Moirai 2.0}
\label{app:official}
\label{app:official-moirai}
%==============================================================================

Our retrained Baseline differs from the official Moirai-2.0-R-Small release due to substantially smaller training compute:

\begin{table}[h]
\centering
\caption{Comparison to official Moirai-2.0-R-Small (HuggingFace release). The gap is expected: the official model was trained with Salesforce's full compute budget.}
\small
\begin{tabular}{lcc}
\toprule
\textbf{Model} & \textbf{Geo Mean MASE} & \textbf{Training} \\
\midrule
Official Moirai-2.0-R-Small (HF) & 0.7280 & Full Salesforce budget \\
Our BL (5-seed mean, 10K) & 0.8617 & 10K steps \\
Our \HD{} (5-seed mean, 10K) & 0.8368 & 10K steps \\
Our BL (5-seed mean, 100K) & 0.8784 & 100K steps \\
Our \HD{} (5-seed mean, 100K) & 0.8443 & 100K steps \\
\bottomrule
\end{tabular}
\end{table}

The gap between our retrained models and the official release is expected and does \emph{not} affect our findings. Our experiments are designed as controlled comparisons: all conditions (Baseline, \HD{}, Zero, Duplicate) use identical training settings, data, and compute. The relative improvement from input preconditioning is the relevant finding, not the absolute MASE level.

\end{document}
